---
title: "Forensic Analysis: TenantThread's Embargo Breach on Project HarborCrest"
author: "Lei Yuhe"
date: "May 31, 2026"
date-modified: "last-modified"
format:
  html:
    toc-depth: 3
    number-sections: true
execute:
  eval: true
  echo: true
  freeze: true
  warning: false
  message: false
---

## Overview

This exercises explores and resolves the [Mini Challenge 1](https://vast-challenge.github.io/2026/MC1.html) of the [Vast Challenge 2026](https://ieeevis.org/year/2026/welcome/).

On June 5, 2046, embargoed information about TenantThread's merger with CivicLoom Realty Partners appeared publicly on FleX, a social media platform, at approximately 5pm, one hour before the agreed 6pm announcement time. This report reconstructs the events and decision-making that led to the breach, using internal agent communications and deliberations from the two weeks preceding the release.

The question we are trying to figure out is:

> **Did TenantThread's team deliberately leak the merger, or did the system simply break down under pressure?**

## Getting Started

### Loading Packages

```{r}
pacman::p_load(tidyverse, jsonlite, lubridate,
  knitr, kableExtra, patchwork, scales, ggrepel, DT
)
```

### Loading and Parsing Data

The dataset is a nested JSON structure with a single top-level key: `rounds`. Each element of `rounds` contains:

-   `hour`: the timestamp for the round
-   `environment_context`: external state (market, social media, events)
-   `communications`: all messages sent during that round, with internal agent states
-   `participants`: declared actions and metadata for each active agent

```{r}
raw <- jsonlite::fromJSON("data/MC1_final_00.json", simplifyVector = FALSE)
rounds <- raw$rounds

get_text <- function(x) {
  if (is.null(x) || length(x) == 0) return(NA_character_)
  return(as.character(x[[1]]))
}

cat("Total rounds:", length(rounds), "\n")
cat("Date range:", get_text(rounds[[1]]$hour), "to", get_text(rounds[[length(rounds)]]$hour))
```

### Creating Tables - Round

::: panel-tabset
#### Table Overview

```{r}
#| echo: false

round_tbl <- map_dfr(seq_along(rounds), function(i) {
  r <- rounds[[i]]
  env <- r$environment_context
  ms <- env$market_snapshot
  t <- ymd_hms(get_text(r$hour))

  tibble(
    round_id = i,
    hour = t,
    stage = case_when(
      t < ymd_hms("2046-06-05 00:00:00") ~ "Pre-crisis",
      t < ymd_hms("2046-06-05 17:00:00") ~ "Crisis (pre-breach)",
      t < ymd_hms("2046-06-05 18:00:00") ~ "Breach window",
      TRUE ~ "Post-embargo"
    ),
    event_headline = get_text(env$event_headline),
    stock_price = parse_number(get_text(ms$stock_price)),
    pct_change = parse_number(get_text(ms$percent_change)),
    market_sentiment = get_text(ms$sentiment),
    n_comms = length(r$communications),
    agents_unavailable = paste(unlist(env$agents_unavailable %||% list()), collapse = ", ")
  )
})

DT::datatable(round_tbl, options = list(pageLength = 5),class = "nowrap")
```

#### Code

**`round_tbl`**: capturing the external environment.

```{r}
#| eval: false

round_tbl <- map_dfr(seq_along(rounds), function(i) {
  r <- rounds[[i]]
  env <- r$environment_context
  ms <- env$market_snapshot
  t <- ymd_hms(get_text(r$hour))

  tibble(
    round_id = i,
    hour = t,
    stage = case_when(
      t < ymd_hms("2046-06-05 00:00:00") ~ "Pre-crisis",
      t < ymd_hms("2046-06-05 17:00:00") ~ "Crisis (pre-breach)",
      t < ymd_hms("2046-06-05 18:00:00") ~ "Breach window",
      TRUE ~ "Post-embargo"
    ),
    event_headline = get_text(env$event_headline),
    stock_price = parse_number(get_text(ms$stock_price)),
    pct_change = parse_number(get_text(ms$percent_change)),
    market_sentiment = get_text(ms$sentiment),
    n_comms = length(r$communications),
    agents_unavailable = paste(unlist(env$agents_unavailable %||% list()), collapse = ", ")
  )
})

DT::datatable(round_tbl, options = list(pageLength = 5),class = "nowrap")
```
:::

### Creating Tables - Communication

::: panel-tabset
#### Table Overview

```{r}
#| echo: false
communication_tbl <- map_dfr(seq_along(rounds), function(i) {
  r <- rounds[[i]]
  env <- r$environment_context
  round_time <- ymd_hms(get_text(r$hour))

  map_dfr(r$communications, function(msg) {
    ist <- msg$internal_state
    t <- ymd_hms(get_text(msg$timestamp))
    content_lc <- str_to_lower(get_text(msg$content))
    channel_text <- get_text(msg$channel)

    # creating a simple data frame, using function&package under tidyverse
    tibble(
      round_id = i,
      hour = round_time,
      timestamp = t,
      stage = case_when(
        t < ymd_hms("2046-06-05 00:00:00") ~ "Pre-crisis",
        t < ymd_hms("2046-06-05 17:00:00") ~ "Crisis (pre-breach)",
        t < ymd_hms("2046-06-05 18:00:00") ~ "Breach window",
        TRUE ~ "Post-embargo"
      ),
      agent_id = get_text(msg$agent_id),
      agent_role = get_text(msg$agent_role),
      channel = channel_text,
      content = get_text(msg$content),
      deliberating = get_text(ist$deliberating),
      reacting = get_text(ist$reacting),
      rationalizing = get_text(ist$rationalizing),
      # Keyword flags
      deal_terms = str_detect(content_lc, "civicloom|harborcrest|residentedge|definitive merger|merger agreement"),
      embargo_terms = str_detect(content_lc, "embargo|6:00|6pm|bilateral release|section 4\\.3"),
      crisis_terms = str_detect(content_lc, "mac|covenant|termination|critical|saltwind"),
      public_channel = channel_text %in% c("official_post", "personal_post", "anonymous_post"),
      anon_channel = channel_text == "anonymous_post",
      side_channel = channel_text == "side_huddle"
    )
  })
})

cat("Total messages:", nrow(communication_tbl))

DT::datatable(communication_tbl, options = list(pageLength = 5), class = "nowrap")
```

#### Code

```{r}
#| eval: false
communication_tbl <- map_dfr(seq_along(rounds), function(i) {
  r <- rounds[[i]]
  env <- r$environment_context
  round_time <- ymd_hms(get_text(r$hour))

  map_dfr(r$communications, function(msg) {
    ist <- msg$internal_state
    t <- ymd_hms(get_text(msg$timestamp))
    content_lc <- str_to_lower(get_text(msg$content))
    channel_text <- get_text(msg$channel)

    # creating a simple data frame, using function&package under tidyverse
    tibble(
      round_id = i,
      hour = round_time,
      timestamp = t,
      stage = case_when(
        t < ymd_hms("2046-06-05 00:00:00") ~ "Pre-crisis",
        t < ymd_hms("2046-06-05 17:00:00") ~ "Crisis (pre-breach)",
        t < ymd_hms("2046-06-05 18:00:00") ~ "Breach window",
        TRUE ~ "Post-embargo"
      ),
      agent_id = get_text(msg$agent_id),
      agent_role = get_text(msg$agent_role),
      channel = channel_text,
      content = get_text(msg$content),
      deliberating = get_text(ist$deliberating),
      reacting = get_text(ist$reacting),
      rationalizing = get_text(ist$rationalizing),
      # Keyword flags
      deal_terms = str_detect(content_lc, "civicloom|harborcrest|residentedge|definitive merger|merger agreement"),
      embargo_terms = str_detect(content_lc, "embargo|6:00|6pm|bilateral release|section 4\\.3"),
      crisis_terms = str_detect(content_lc, "mac|covenant|termination|critical|saltwind"),
      public_channel = channel_text %in% c("official_post", "personal_post", "anonymous_post"),
      anon_channel = channel_text == "anonymous_post",
      side_channel = channel_text == "side_huddle"
    )
  })
})

cat("Total messages:", nrow(communication_tbl))

DT::datatable(communication_tbl, options = list(pageLength = 5), class = "nowrap")
```
:::

### Creating Tables - Participant (agents)

::: panel-tabset
#### Table Overview

```{r}
#| echo: false
participant_tbl <- map_dfr(seq_along(rounds), function(i) {
  r <- rounds[[i]]
  round_time <- ymd_hms(get_text(r$hour))

  map_dfr(r$participants %||% list(), function(p) {
    tibble(
      round_id = i,
      hour = round_time,
      agent_id = get_text(p$agent_id),
      declared_action = get_text(p$declared_action)
    )
  })
})

DT::datatable(participant_tbl, options = list(pageLength = 5),class = "nowrap")
```

#### Code

```{r}
#| eval: false
participant_tbl <- map_dfr(seq_along(rounds), function(i) {
  r <- rounds[[i]]
  round_time <- ymd_hms(get_text(r$hour))

  map_dfr(r$participants %||% list(), function(p) {
    tibble(
      round_id = i,
      hour = round_time,
      agent_id = get_text(p$agent_id),
      declared_action = get_text(p$declared_action)
    )
  })
})

DT::datatable(participant_tbl, options = list(pageLength = 5),class = "nowrap")
```
:::

### Defining the Visual Theme

```{r}
theme_forensic <- function() {
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 13, margin = margin(b = 6)),
      plot.subtitle = element_text(size = 10, color = "grey40", margin = margin(b = 10)),
      plot.caption = element_text(size = 8, color = "grey55", hjust = 0),
      axis.title = element_text(size = 10, color = "grey30"),
      axis.text = element_text(size = 9,  color = "grey30"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey92"),
      legend.position = "top",
      legend.title = element_text(size = 9, face = "bold"),
      legend.text = element_text(size = 8),
      strip.text = element_text(size = 9, face = "bold")
    )
}

# A minimal two-colour palette: dark for primary, mid-grey for secondary
COL_PRIMARY <- "#2b2b2b"
COL_SECONDARY <- "#888888"
COL_ALERT <- "#c0392b"
COL_STAGE <- c(
  "Pre-crisis" = "#bbbbbb",
  "Crisis (pre-breach)" = "#666666",
  "Breach window" = "#c0392b",
  "Post-embargo" = "#2b6cb0"
)
```

------------------------------------------------------------------------

## Task 1. Key Events and Relationships Leading to the Release

### The Two-Week Window

```{r}
#| label: fig-window
#| fig-cap: "Stock price trajectory and message activity across all 23 rounds. Each round is one dot where size represents communication volume. Vertical bands mark the four narrative stages."
#| fig-height: 5
#| fig-width: 10
#| code-fold: true

# Fix the two anomalous stock prices that appear in market_snapshot
# (the $18 and $180 values are from a corrupted field on Jun 5 hourly rounds;
#  agent messages quote intraday prices of $28.70 → $25.80 → $33.05)
stock_clean <- round_tbl %>%
  mutate(stock_price = case_when(
    hour == ymd_hms("2046-06-05 09:00:00") ~ 28.70,
    hour == ymd_hms("2046-06-05 10:00:00") ~ 27.80,
    hour == ymd_hms("2046-06-05 11:00:00") ~ 27.20,
    hour == ymd_hms("2046-06-05 12:00:00") ~ 27.20,
    hour == ymd_hms("2046-06-05 13:00:00") ~ 26.90,
    hour == ymd_hms("2046-06-05 14:00:00") ~ 26.40,
    hour == ymd_hms("2046-06-05 15:00:00") ~ 25.80,
    hour == ymd_hms("2046-06-05 16:00:00") ~ 26.20,
    hour == ymd_hms("2046-06-05 17:00:00") ~ 26.80,
    hour == ymd_hms("2046-06-05 18:00:00") ~ 33.05,
    TRUE ~ stock_price
  ))

# Shade rectangles for each stage
stage_bands <- tribble(
  ~xmin,                          ~xmax,                          ~stage,
  ymd_hms("2046-05-17 00:00:00"), ymd_hms("2046-06-05 00:00:00"), "Pre-crisis",
  ymd_hms("2046-06-05 00:00:00"), ymd_hms("2046-06-05 17:00:00"), "Crisis (pre-breach)",
  ymd_hms("2046-06-05 17:00:00"), ymd_hms("2046-06-05 18:00:00"), "Breach window",
  ymd_hms("2046-06-05 18:00:00"), ymd_hms("2046-06-06 00:00:00"), "Post-embargo"
)


ggplot() +
  geom_rect(data = stage_bands,
            aes(xmin = xmin, xmax = xmax,
                ymin = -Inf, ymax = Inf, fill = stage),
            alpha = 0.08, inherit.aes = FALSE) +
  scale_fill_manual(values = COL_STAGE, name = "Stage") +
  geom_line(data = stock_clean %>% filter(!is.na(stock_price)),
            aes(x = hour, y = stock_price),
            colour = COL_PRIMARY, linewidth = 0.8) +
  geom_point(data = stock_clean %>% filter(!is.na(stock_price)),
             aes(x = hour, y = stock_price, size = n_comms),
             colour = COL_PRIMARY, alpha = 0.7) +
  geom_vline(xintercept = ymd_hms("2046-06-05 17:25:00"),
             colour = COL_ALERT, linetype = "dashed", linewidth = 0.8) +
  scale_size_continuous(range = c(2, 8), name = "Messages\nper round") +
  scale_x_datetime(date_labels = "%b %d", date_breaks = "4 days") +
  scale_y_continuous(labels = dollar_format(prefix = "$")) +
  labs(title    = "TenantThread Stock Price — May 17 to Jun 5, 2046",
       subtitle = "Point size = messages per round. Dashed red = breach moment (17:25).",
       x = NULL, y = "Stock price (USD)") +
  theme_forensic()
```

The breach did not emerge from a single bad decision. It was the endpoint of a 20-day escalation in which three forces converged: a worsening public narrative (the SaltWind expose), a secret corporate transaction (Project HarborCrest) known to senior agents since week one, and a compliance system (Judge-Agent) installed mid-crisis with critical blind spots.

### The Breach Sequence

The actual breach comprised four rapid events inside a 30-minute window. For the first time all day, agents acted *without* waiting for Judge-Agent's sign-off.

```{r}
#| label: tbl-breach-sequence
#| tbl-cap: "The four public posts that constituted the breach, all on personal or anonymous channels the Judge had no authority to review."
#| code-fold: true


breach_msgs <- communication_tbl %>%
  filter(stage == "Breach window", public_channel, deal_terms) %>%
  select(timestamp, agent_id, channel, content) %>%
  mutate(
    content  = str_trunc(content, 120),
    agent_id = str_replace_all(agent_id, "_agent", "")
  ) %>%
  arrange(timestamp)

kable(breach_msgs,
      col.names = c("Timestamp", "Agent", "Channel", "Content (truncated)"),
      align     = c("l","l","l","l")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = TRUE, font_size = 13) %>%
  row_spec(0, bold = TRUE)
```



### Causal Relationships and Decision Points

Each row below is one contributing event. The **Causal link** column describes how it enabled the next step, **actor** identifies who owned it, **gap** identifies a structural failure that prevented remediation.


```{r}
#| label: tbl-causal-chain
#| tbl-cap: "Causal chain from first structural failure (May 22) to embargo breach (Jun 5 17:25). Rows are in chronological order"
#| code-fold: true

causal_tbl <- tribble(
  ~Date,            ~Event,                                        ~Actor,              ~Causal_link,                                                                                     ~Gap,                                                        ~Stage,
  "May 22",         "Shadow channel first used off-record",        "Senior team",       "Sets precedent for coordination outside the Judge's view",                                       "No oversight body yet",                                     "Pre-crisis",
  "May 29",         "@ElenaMarquez tagged on personal post",       "Social-Manager",    "Near-breach; reveals personal posts carry embargo risk",                                         "Judge installed but personal posts explicitly excluded",  "Pre-crisis",
  "May 30",         "Judge installed",                             "Legal + CEO",       "Adds compliance layer, but remit excludes personal posts and side_huddle",                       "Structural blind spot baked in from day one",               "Pre-crisis",
  "May 31",         "SaltWind Piece #1 published",                 "External (press)",  "Raises public pressure; pushes team toward defensive comms",                                    "—",                                                         "Pre-crisis",
  "Jun 1",          "Judge clears response language",              "Judge",             "Establishes pattern: Judge approves official posts; personal posts remain self-approved",        "—",                                                         "Pre-crisis",
  "Jun 4",          "SaltWind Piece #2 (risk score 9.4/10)",       "External (press)",  "CRITICAL posture declared; merger announcement becomes urgent exit strategy",                   "—",                                                         "Pre-crisis",
  "Jun 5, 09:00",  "OceanCrunch 'Silence' piece; stock −8%",     "External",          "MAC clause exposure becomes imminent; Legal switches to survival mode",                         "—",                                                         "Crisis (pre-breach)",
  "Jun 5, 13:00",  "CivicLoom invokes Section 4.2(c)",           "CivicLoom legal",   "Contractual threat that deal collapses if stock keeps falling",                                    "—",                                                         "Crisis (pre-breach)",
  "Jun 5, 15:08",  "Judge issues COMPLIANCE_WARNING",            "Judge",             "Acknowledges ceiling reached, warns against further forward-looking language",                  "Warning is advisory only, not a blocking mechanism",        "Crisis (pre-breach)",
  "Jun 5, 17:00",  "SaltWind independently publishes merger",    "External (press)",  "Legal reframes embargo as moot: 'third-party sourced, not our breach'",                        "No automatic re-verification with CivicLoom required",     "Breach window",
  "Jun 5, 17:19",  "Legal declares verbal consent, issues 'GO'", "Legal",             "Triggers full execution cascade (PR-Intern, Social-Manager, Intern all pre-staged and ready)", "Verbal consent unverified; Judge not re-consulted",         "Breach window",
  "Jun 5, 17:25",  "Legal posts CivicLoom name on personal account", "Legal",         "Merger named publicly 35 minutes before the 6 PM embargo lift",                   "Personal post channel was never within the Judge's scope",  "Breach window"
)

stage_bg <- c(
  "Pre-crisis"          = "#f0f0f0",
  "Crisis (pre-breach)" = "#f5e6c8",
  "Breach window"       = "#fde8e8"
)

causal_tbl %>%
  select(Date, Event, Actor, Causal_link, Gap, Stage) %>%
  kable(
    col.names = c("Date", "Event", "Actor", "Causal link to next step",
                  "Structural gap / why no remediation", "Stage"),
    align     = c("l", "l", "l", "l", "l", "l"),
    escape    = FALSE
  ) %>%
  kable_styling(
    bootstrap_options = c("hover", "condensed"),
    full_width        = TRUE,
    font_size         = 12
  ) %>%
  column_spec(1, bold = TRUE, width = "7em") %>%
  column_spec(2, width = "14em") %>%
  column_spec(3, width = "9em",  color = "grey30") %>%
  column_spec(4, width = "18em") %>%
  column_spec(5, width = "16em", color = "#c0392b") %>%
  column_spec(6, width = "9em") %>%
  row_spec(1:6,   background = stage_bg["Pre-crisis"]) %>%
  row_spec(7:9,   background = stage_bg["Crisis (pre-breach)"]) %>%
  row_spec(10:12, background = stage_bg["Breach window"]) %>%
  row_spec(12, bold = TRUE) %>%
  row_spec(0,  bold = TRUE)
```


## Task 2. Typical Behaviour vs Breach Behaviour

### Communication Volume by Stage

The most immediate signal of abnormal behaviour is sheer volume. On the crisis day, Legal-Agent and Social-Manager-Agent each sent more messages than across the entire preceding two weeks.

```{r}
#| label: fig-volume-stage
#| fig-cap: "Total messages per agent across each stage. Crisis-day volume (Crisis + Breach) is 10–15× the pre-crisis daily average."
#| fig-height: 4.5
#| code-fold: true

vol_by_agent_stage <- communication_tbl %>%
  mutate(agent_short = str_remove(agent_id, "_agent")) %>%
  count(agent_short, stage)

# Order agents by total volume
agent_order <- vol_by_agent_stage %>%
  group_by(agent_short) %>% summarise(tot = sum(n), .groups = "drop") %>%
  arrange(tot) %>% pull(agent_short)

vol_by_agent_stage %>%
  mutate(agent_short = factor(agent_short, levels = agent_order),
         stage       = factor(stage, levels = names(COL_STAGE))) %>%
  ggplot(aes(x = n, y = agent_short, fill = stage)) +
  geom_col(width = 0.7, alpha = 0.88) +
  scale_fill_manual(values = COL_STAGE, name = "Stage") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(title    = "Total Messages per Agent by Stage",
       subtitle = "Legal and Social-Manager dominate the crisis and breach stages.",
       x = "Message count", y = NULL) +
  theme_forensic()
```

### Channel Behaviour: Public Posting Patterns

Agents have a clear pre-crisis posting frequency: 1–2 public posts per day, all via `official_post`. On June 5, three new posting patterns appeared for the first time.

```{r}
#| label: fig-channel-shift
#| fig-cap: "Public-facing post count by channel type and stage. The anonymous and personal channels were used almost exclusively on the crisis day."
#| fig-height: 4
#| code-fold: true

pub_channel <- communication_tbl %>%
  filter(public_channel) %>%
  mutate(stage = factor(stage, levels = names(COL_STAGE))) %>%
  count(stage, channel)

ggplot(pub_channel, aes(x = n, y = channel, fill = stage)) +
  geom_col(position = "dodge", width = 0.65, alpha = 0.88) +
  scale_fill_manual(values = COL_STAGE, name = "Stage") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(title    = "Public Posts by Channel and Stage",
       subtitle = "Anonymous posts and high-volume personal posts are exclusive to June 5.",
       x = "Post count", y = NULL) +
  theme_forensic()
```

### Side-Huddle: The Coordination Layer without Judge

The `side_huddle` channel is a private, off-record channel invisible to the Judge. It was first activated on May 22 for merger-related discussion. Its usage exploded on the crisis day with 99 messages, all unreviewed by the judge. 

```{r}
#| label: fig-sidehuddle
#| fig-cap: "Side-huddle messages per day. The channel was dormant for most of the pre-crisis period, then erupted on June 5."
#| fig-height: 4
#| fig-width: 8
#| code-fold: true

side_daily <- communication_tbl %>%
  filter(side_channel) %>%
  mutate(date = as_date(timestamp)) %>%
  count(date, agent_id) %>%
  mutate(agent_short = str_remove(agent_id, "_agent"))

# Fill in zero days
all_days <- tibble(date = seq(as_date("2046-05-17"),
                               as_date("2046-06-05"), by = "day"))
side_total <- side_daily %>%
  group_by(date) %>% summarise(n = sum(n), .groups = "drop") %>%
  right_join(all_days, by = "date") %>%
  replace_na(list(n = 0)) %>%
  mutate(is_crisis = date == as_date("2046-06-05"))

ggplot(side_total, aes(x = date, y = n, fill = is_crisis)) +
  geom_col(width = 0.8, alpha = 0.88) +
  scale_fill_manual(values = c("FALSE" = COL_SECONDARY, "TRUE" = COL_ALERT),
                    labels = c("FALSE" = "Normal day", "TRUE" = "Crisis day"),
                    name   = NULL) +
  annotate("text", x = as_date("2046-05-22"), y = 8,
           label = "First use\n(May 22)",
           colour = COL_SECONDARY, size = 3, hjust = 1.5) +
  annotate("text", x = as_date("2046-06-05"), y = 103,
           label = "99 msgs \nJudge cannot see",
           colour = COL_ALERT, size = 3, hjust = 1.5) +
  scale_x_date(date_labels = "%b %d", date_breaks = "4 days") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(title    = "Side-Huddle Message Volume by Day",
       subtitle = "The Judge's mandate never covered this channel.",
       x = NULL, y = "Messages") +
  theme_forensic()
```

### Behavioural Summary Table

```{r}
#| label: tbl-behaviour-compare
#| tbl-cap: "Baseline vs crisis-day behaviour for each agent across four dimensions."
#| code-fold: true
#| fig-width: 10

# Pre-crisis daily average (across 13 pre-crisis rounds)
n_pre_rounds <- round_tbl %>% filter(stage == "Pre-crisis") %>% nrow()

pre_avg <- communication_tbl %>%
  filter(stage == "Pre-crisis") %>%
  group_by(agent_id) %>%
  summarise(
    avg_msgs   = round(n() / n_pre_rounds, 1),
    avg_pub    = round(sum(public_channel, na.rm = TRUE) / n_pre_rounds, 1),
    avg_side   = round(sum(side_channel,   na.rm = TRUE) / n_pre_rounds, 1),
    avg_anon   = round(sum(anon_channel,   na.rm = TRUE) / n_pre_rounds, 1),
    .groups = "drop"
  ) %>%
  rename_with(~ str_replace(., "avg_", "pre_"), -agent_id)

crisis_day <- communication_tbl %>%
  filter(stage %in% c("Crisis (pre-breach)", "Breach window")) %>%
  group_by(agent_id) %>%
  summarise(
    crisis_msgs = n(),
    crisis_pub  = sum(public_channel, na.rm = TRUE),
    crisis_side = sum(side_channel,   na.rm = TRUE),
    crisis_anon = sum(anon_channel,   na.rm = TRUE),
    .groups = "drop"
  )

behaviour_tbl <- full_join(pre_avg, crisis_day, by = "agent_id") %>%
  replace_na(list(pre_msgs=0, pre_pub=0, pre_side=0, pre_anon=0,
                  crisis_msgs=0, crisis_pub=0, crisis_side=0, crisis_anon=0)) %>%
  mutate(agent_id = str_remove(agent_id, "_agent")) %>%
  arrange(desc(crisis_msgs))

kable(behaviour_tbl,
      col.names = c("Agent", "Msgs/day", "Public", "Side", "Anon", "Msgs", "Public", "Side", "Anon"),
      align = "lrrrrrrrr") %>%
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = TRUE, font_size = 12) %>%
  column_spec(6, bold = TRUE, color = "blue",
              background = spec_color(behaviour_tbl$crisis_msgs,
                                      palette = "Greys", end = 0.7)) %>%
  add_header_above(c(" " = 1, "Pre-crisis baseline" = 4,
                     "Crisis day total" = 4))
```

---

## Task 3. Leading Indicators and Prior Near-Breaches 

### Were There Leading Indicators?



```{r}
#| label: fig-leading-indicators
#| fig-cap: "Three leading indicator series plotted together. All three began escalating before the formal crisis began. Dashed vertical lines mark the @Elena incident (May 29), SaltWind journal publishment #2 (Jun 4), and the breach (Jun 5 17:25)."
#| fig-height: 10
#| code-fold: true

daily_signals <- communication_tbl %>%
  mutate(date = as_date(timestamp)) %>%
  group_by(date) %>%
  summarise(
    total_msgs   = n(),
    side_msgs    = sum(side_channel, na.rm = TRUE),
    deal_msgs    = sum(deal_terms,   na.rm = TRUE),
    delib_chars  = sum(nchar(replace_na(deliberating, "")), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  right_join(tibble(date = seq(as_date("2046-05-17"),
                                as_date("2046-06-05"), by = "day")),
             by = "date") %>%
  replace_na(list(total_msgs=0, side_msgs=0, deal_msgs=0, delib_chars=0))

key_vlines <- as_date(c("2046-05-29", "2046-06-04", "2046-06-05"))
vline_labels <- c("@Elena\nincident", "SaltWind #2\n(9.4/10)", "Breach\nday")

add_vlines <- function(p) {
  p +
    geom_vline(xintercept = key_vlines, linetype = "dotted",
               colour = "Blue", linewidth = 0.6) +
    annotate("text", x = key_vlines, y = Inf,
             label = vline_labels,
             vjust = 0.9, hjust = 1.2, size = 2,
             colour = "blue", lineheight = 0.85)
}

p1 <- ggplot(daily_signals, aes(x = date, y = side_msgs)) +
  geom_col(fill = COL_PRIMARY, alpha = 0.75, width = 0.8) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(title = "A. Side-huddle messages per day",
       subtitle = "First use May 22; explodes on Jun 5. Judge never monitored this channel.",
       x = NULL, y = "Messages") +
  theme_forensic()
p1 <- add_vlines(p1)

p2 <- ggplot(daily_signals, aes(x = date, y = deal_msgs)) +
  geom_col(fill = COL_SECONDARY, alpha = 0.8, width = 0.8) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(title = "B. Deal-term messages per day",
       subtitle = "CivicLoom / HarborCrest references appear in internal messages weeks before the breach.",
       x = NULL, y = "Messages") +
  theme_forensic()
p2 <- add_vlines(p2)

p3 <- ggplot(daily_signals, aes(x = date, y = delib_chars / 1000)) +
  geom_col(fill = COL_ALERT, alpha = 0.7, width = 0.8) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)),
                     labels = label_comma(suffix = "k")) +
  labs(title = "C. Internal deliberation length per day (characters)",
       subtitle = "Agents' private reasoning about the merger escalates steadily from May 22 onward.",
       x = NULL, y = "Characters (thousands)") +
  theme_forensic()
p3 <- add_vlines(p3)

p1 / p2 / p3 +
  plot_annotation(
    title   = "Three Leading Indicators That Preceded the Breach",
    caption = "Dotted lines: @Elena incident (May 29) | SaltWind #2 (Jun 4) | Breach day (Jun 5)"
  )
```

### The @Elena Incident: A Direct Preview of the Breach Mechanism

On May 29, Social-Manager-Agent posted to their *personal* Flex account: *"So energized by the PropTech leaders shaping this space, @ElenaMarquez @AjayTT … Big things coming!"* inadvertently tagging the CEO of CivicLoom. The post was live for 14 minutes. A CivicLoom employee account liked it before deletion.

The response was to install the Judge. But the Judge's oversight only covers for Comms_huddle and **explicitly excluded personal posts**, the exact channel later used for the breach.

```{r}
#| label: fig-elena-response
#| fig-cap: "Message traffic on May 29 by channel. The incident generated a spike in 1-on-1 coordination but no change to personal-post policy."
#| fig-height: 3.5
#| fig-width: 8
#| code-fold: true

elena_day <- communication_tbl %>%
  filter(as_date(timestamp) == as_date("2046-05-29")) %>%
  mutate(channel_clean = recode(channel,
    comms_huddle    = "Comms huddle",
    one_on_one_chat = "1-on-1 chat",
    personal_post   = "Personal post",
    side_huddle     = "Side huddle"
  )) %>%
  count(channel_clean) %>%
  arrange(desc(n))

ggplot(elena_day, aes(x = n, y = reorder(channel_clean, n))) +
  geom_col(fill = COL_PRIMARY, alpha = 0.8, width = 0.6) +
  geom_text(aes(label = n), hjust = -0.3, size = 3.5, colour = COL_PRIMARY) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title    = "May 29: Channel Usage on the Day of the @Elena Incident",
       subtitle = "Personal post that nearly breached the embargo prompted 1-on-1 damage control,\nbut no change to personal-post oversight policy.",
       x = "Messages", y = NULL) +
  theme_forensic()
```

### Prior Occasions Where Behaviour Differed from Expected

The table below lists three occasions, where agents acted outside their expected patterns. None triggered a structural fix.

```{r}
#| label: tbl-near-breaches
#| tbl-cap: "Prior occasions where agent behaviour deviated from expected norms and previewed the breach mechanism."
#| code-fold: true

near_breach_tbl <- tribble(
  ~Date,       ~Incident,                        ~Expected,                          ~Actual,                                              ~Why_no_action,
  "May 22",    "Side-huddle activated",           "Strategy discussed in comms_huddle", "Senior agents move off-record to discuss merger",  "No oversight body yet; not flagged as a violation",
  "May 23",    "Merger inferred by senior agents","Only Ajay to know deal details",    "All senior agents privately confirm the deal",      "CEO briefed each privately; embargo treated as informal",
  "May 29",    "@Elena personal post",            "Public posts routed through PR",    "Social-Manager posts merger-adjacent content\npublicly, tagging CivicLoom CEO", "Post deleted in 14 min; Judge installed, but personal posts explicitly excluded",
  "Jun 5 10am","Intern leaks 'CivicLoom 6 PM'",   "Interns not briefed on merger",    "General intern mentions embargo timeline in hallway; PR agent contains it informally", "Treated as minor containment; no re-assessment of information perimeter"
)

kable(near_breach_tbl,
      col.names = c("Date","Incident","Expected behaviour",
                    "Actual behaviour","Why no remediation"),
      align = c("l","l","l","l","l")) %>%
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = TRUE, font_size = 12) %>%
  column_spec(4, italic = TRUE) %>%
  column_spec(5, color = COL_ALERT) %>%
  row_spec(0, bold = TRUE)
```

### Why the Judge Failed to Prevent the Breach

The Judge's own message on May 30 defines its remit: *"For posts that are clearly routine … I'm fine with you approving those yourself. I want to reserve my review for anything that names leadership, mentions partnerships, uses forward-looking language, or touches on data practices."*

This language excluded personal posts entirely. The four breach posts were all on `personal_post` or `anonymous_post`, channels the Judge had never been asked to monitor.

```{r}
#| label: fig-judge-scope
#| fig-cap: "Judge-Agent message volume and decisions by date. The Judge issued only one COMPLIANCE_WARNING on the day of the breach and it applied to a comms_huddle statement, not to the personal-post channel used for the actual breach."
#| fig-height: 3.5
#| fig-width: 7
#| code-fold: true

judge_activity <- communication_tbl %>%
  filter(agent_id == "judge_agent") %>%
  mutate(date = as_date(timestamp),
         decision = case_when(
           str_detect(str_to_lower(content), "cleared|approved|defensible|execute") ~ "Approve",
           str_detect(str_to_lower(content), "compliance_warning|flag|hold")        ~ "Flag / warn",
           TRUE                                                                      ~ "Observe"
         ))

ggplot(judge_activity, aes(x = date, fill = decision)) +
  geom_bar(width = 0.7, alpha = 0.88) +
  scale_fill_manual(
    values = c("Approve" = "#2b6cb0", "Flag / warn" = COL_ALERT,
               "Observe" = COL_SECONDARY),
    name = "Decision type"
  ) +
  scale_x_date(date_labels = "%b %d", date_breaks = "3 days") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(title    = "Judge-Agent Activity by Date",
       subtitle = "Judge issued one warning on Jun 4 in comms_huddle, not on the personal-post channel.",
       x = NULL, y = "Messages") +
  theme_forensic()
```

---

## Conclusion

The breach was neither a premeditated leak nor a pure system accident. It was **rationalised action under extreme pressure**, enabled by three structural gaps.

1. No direct evidence that the breach was deliberate. No message expressed the intent to breach the embargo before 5 Jun, 6pm. 

2. The Judge was partially effective. It only reviews the comms_huddle and official posts with reasonable judgement. It was never given authority to oversee personal posts or side_huddle, the two channels where the breach happened. 

3. Side-huddle activation (May 22), @Elena in a private post (May 29), intern hallway leak (Jun 5. 10am), and escalating anonymous post usage on June 5 from 9am all previewed the breach. None triggered a change to personal-post policy.



---









