---
title: "TenantThread Embargo Breach: Forensic Communication Analysis"
author: "Lei Yuhe"
date: "May 31, 2026"
date-modified: "last-modified"
format:
  html:
    toc-depth: 3
    number-sections: true
execute:
  eval: true
  echo: true
  freeze: true
  warning: false
  message: false
---

## Overview

This exercises explores and resolves the [Mini Challenge 1](https://vast-challenge.github.io/2026/MC1.html) of the [Vast Challenge 2026](https://ieeevis.org/year/2026/welcome/).

On June 5, 2046, embargoed information about TenantThread's merger with CivicLoom Realty Partners appeared publicly on FleX, a social media platform, at approximately 5pm, one hour before the agreed 6pm announcement time. This report reconstructs the events and decision-making that led to the breach, using internal agent communications and deliberations from the two weeks preceding the release.

The question we are trying to figure out is:

> **Did TenantThread's team deliberately leak the merger, or did the system simply break down under pressure?**

## Getting Started

### Loading Packages

```{r}
pacman::p_load(tidyverse, jsonlite, lubridate,
  knitr, kableExtra, patchwork, scales, ggrepel
)
```
### Loading and Parsing Data

The dataset is a nested JSON structure with a single top-level key: `rounds`. Each element of `rounds` contains:

- `hour`: the timestamp for the round
- `environment_context`: external state (market, social media, events)
- `communications`: all messages sent during that round, with internal agent states
- `participants`: declared actions and metadata for each active agent

```{r}
raw <- jsonlite::fromJSON("data/MC1_final_00.json", simplifyVector = FALSE)
rounds <- raw$rounds

get_text <- function(x) {
  if (is.null(x) || length(x) == 0) return(NA_character_)
  return(as.character(x[[1]]))
}

cat("Total rounds:", length(rounds), "\n")
cat("Date range:", get_text(rounds[[1]]$hour), "to", get_text(rounds[[length(rounds)]]$hour))
```
### Creating Tables - Round

::: {.panel-tabset}

#### Table Overview

```{r}
#| echo: false

round_tbl <- map_dfr(seq_along(rounds), function(i) {
  r <- rounds[[i]]
  env <- r$environment_context
  ms <- env$market_snapshot
  t <- ymd_hms(get_text(r$hour))

  tibble(
    round_id = i,
    hour = t,
    stage = case_when(
      t < ymd_hms("2046-06-05 00:00:00") ~ "Pre-crisis",
      t < ymd_hms("2046-06-05 17:00:00") ~ "Crisis (pre-breach)",
      t < ymd_hms("2046-06-05 18:00:00") ~ "Breach window",
      TRUE ~ "Post-embargo"
    ),
    event_headline = get_text(env$event_headline),
    stock_price = parse_number(get_text(ms$stock_price)),
    pct_change = parse_number(get_text(ms$percent_change)),
    market_sentiment = get_text(ms$sentiment),
    n_comms = length(r$communications),
    agents_unavailable = paste(unlist(env$agents_unavailable %||% list()), collapse = ", ")
  )
})

DT::datatable(round_tbl, options = list(pageLength = 5),class = "nowrap")
```

#### Code

**`round_tbl`**: capturing the external environment.

```{r}
#| eval: false

round_tbl <- map_dfr(seq_along(rounds), function(i) {
  r <- rounds[[i]]
  env <- r$environment_context
  ms <- env$market_snapshot
  t <- ymd_hms(get_text(r$hour))

  tibble(
    round_id = i,
    hour = t,
    stage = case_when(
      t < ymd_hms("2046-06-05 00:00:00") ~ "Pre-crisis",
      t < ymd_hms("2046-06-05 17:00:00") ~ "Crisis (pre-breach)",
      t < ymd_hms("2046-06-05 18:00:00") ~ "Breach window",
      TRUE ~ "Post-embargo"
    ),
    event_headline = get_text(env$event_headline),
    stock_price = parse_number(get_text(ms$stock_price)),
    pct_change = parse_number(get_text(ms$percent_change)),
    market_sentiment = get_text(ms$sentiment),
    n_comms = length(r$communications),
    agents_unavailable = paste(unlist(env$agents_unavailable %||% list()), collapse = ", ")
  )
})

DT::datatable(round_tbl, options = list(pageLength = 5),class = "nowrap")
```

:::

### Creating Tables - Communication

::: {.panel-tabset}

#### Table Overview 
```{r}
#| echo: false
communication_tbl <- map_dfr(seq_along(rounds), function(i) {
  r <- rounds[[i]]
  env <- r$environment_context
  round_time <- ymd_hms(get_text(r$hour))

  map_dfr(r$communications, function(msg) {
    ist <- msg$internal_state
    t <- ymd_hms(get_text(msg$timestamp))
    content_lc <- str_to_lower(get_text(msg$content))
    channel_text <- get_text(msg$channel)

    # creating a simple data frame, using function&package under tidyverse
    tibble(
      round_id = i,
      hour = round_time,
      timestamp = t,
      stage = case_when(
        t < ymd_hms("2046-06-05 00:00:00") ~ "Pre-crisis",
        t < ymd_hms("2046-06-05 17:00:00") ~ "Crisis (pre-breach)",
        t < ymd_hms("2046-06-05 18:00:00") ~ "Breach window",
        TRUE ~ "Post-embargo"
      ),
      agent_id = get_text(msg$agent_id),
      agent_role = get_text(msg$agent_role),
      channel = channel_text,
      content = get_text(msg$content),
      deliberating = get_text(ist$deliberating),
      reacting = get_text(ist$reacting),
      rationalizing = get_text(ist$rationalizing),
      # Keyword flags
      deal_terms = str_detect(content_lc, "civicloom|harborcrest|residentedge|definitive merger|merger agreement"),
      embargo_terms = str_detect(content_lc, "embargo|6:00|6pm|bilateral release|section 4\\.3"),
      crisis_terms = str_detect(content_lc, "mac|covenant|termination|critical|saltwind"),
      public_channel = channel_text %in% c("official_post", "personal_post", "anonymous_post"),
      anon_channel = channel_text == "anonymous_post",
      side_channel = channel_text == "side_huddle"
    )
  })
})

cat("Total messages:", nrow(communication_tbl))

DT::datatable(communication_tbl, options = list(pageLength = 5), class = "nowrap")
```


#### Code 
```{r}
#| eval: false
communication_tbl <- map_dfr(seq_along(rounds), function(i) {
  r <- rounds[[i]]
  env <- r$environment_context
  round_time <- ymd_hms(get_text(r$hour))

  map_dfr(r$communications, function(msg) {
    ist <- msg$internal_state
    t <- ymd_hms(get_text(msg$timestamp))
    content_lc <- str_to_lower(get_text(msg$content))
    channel_text <- get_text(msg$channel)

    # creating a simple data frame, using function&package under tidyverse
    tibble(
      round_id = i,
      hour = round_time,
      timestamp = t,
      stage = case_when(
        t < ymd_hms("2046-06-05 00:00:00") ~ "Pre-crisis",
        t < ymd_hms("2046-06-05 17:00:00") ~ "Crisis (pre-breach)",
        t < ymd_hms("2046-06-05 18:00:00") ~ "Breach window",
        TRUE ~ "Post-embargo"
      ),
      agent_id = get_text(msg$agent_id),
      agent_role = get_text(msg$agent_role),
      channel = channel_text,
      content = get_text(msg$content),
      deliberating = get_text(ist$deliberating),
      reacting = get_text(ist$reacting),
      rationalizing = get_text(ist$rationalizing),
      # Keyword flags
      deal_terms = str_detect(content_lc, "civicloom|harborcrest|residentedge|definitive merger|merger agreement"),
      embargo_terms = str_detect(content_lc, "embargo|6:00|6pm|bilateral release|section 4\\.3"),
      crisis_terms = str_detect(content_lc, "mac|covenant|termination|critical|saltwind"),
      public_channel = channel_text %in% c("official_post", "personal_post", "anonymous_post"),
      anon_channel = channel_text == "anonymous_post",
      side_channel = channel_text == "side_huddle"
    )
  })
})

cat("Total messages:", nrow(communication_tbl))

DT::datatable(communication_tbl, options = list(pageLength = 5), class = "nowrap")
```
:::


### Creating Tables - Participant (agents)


::: {.panel-tabset}

#### Table Overview 

```{r}
#| echo: false
participant_tbl <- map_dfr(seq_along(rounds), function(i) {
  r <- rounds[[i]]
  round_time <- ymd_hms(get_text(r$hour))

  map_dfr(r$participants %||% list(), function(p) {
    tibble(
      round_id = i,
      hour = round_time,
      agent_id = get_text(p$agent_id),
      declared_action = get_text(p$declared_action)
    )
  })
})

DT::datatable(participant_tbl, options = list(pageLength = 5),class = "nowrap")
```


#### Code

```{r}
#| eval: false
participant_tbl <- map_dfr(seq_along(rounds), function(i) {
  r <- rounds[[i]]
  round_time <- ymd_hms(get_text(r$hour))

  map_dfr(r$participants %||% list(), function(p) {
    tibble(
      round_id = i,
      hour = round_time,
      agent_id = get_text(p$agent_id),
      declared_action = get_text(p$declared_action)
    )
  })
})

DT::datatable(participant_tbl, options = list(pageLength = 5),class = "nowrap")
```
:::

### Defining the Visual Theme

```{r}
theme_forensic <- function() {
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 13, margin = margin(b = 6)),
      plot.subtitle = element_text(size = 10, color = "grey40", margin = margin(b = 10)),
      plot.caption = element_text(size = 8, color = "grey55", hjust = 0),
      axis.title = element_text(size = 10, color = "grey30"),
      axis.text = element_text(size = 9,  color = "grey30"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey92"),
      legend.position = "top",
      legend.title = element_text(size = 9, face = "bold"),
      legend.text = element_text(size = 8),
      strip.text = element_text(size = 9, face = "bold")
    )
}

# A minimal two-colour palette: dark for primary, mid-grey for secondary
COL_PRIMARY <- "#2b2b2b"
COL_SECONDARY <- "#888888"
COL_ALERT <- "#c0392b"
COL_STAGE <- c(
  "Pre-crisis" = "#bbbbbb",
  "Crisis (pre-breach)" = "#666666",
  "Breach window" = "#c0392b",
  "Post-embargo" = "#2b6cb0"
)
```

---

## Task 1. Key Events and Relationships Leading to the Release

### Background

TenantThread actually operated under a dual crisis:

1. SaltWind Journal's ongoing media report on algorithmic tenant scoring, with the hashtag `#AlgorithmicEviction` gaining traction. Internal documentation showed that the company's "anonymized" data could be reverse-engineered to identify individual tenants at smaller properties.

2. A confidential merger agreement with CivicLoom Realty Partners, named "Project HarborCrest", under strict information embargo until 6PM on 5th June. However, the embargo was broken 1 hour before the agreed release time. CivicLoom could walk away if TenantThread's market sentiment or stock price deteriorated sufficiently.

These two crises were not independent. The scandal directly threatened the conditions under which the merger could close.

### Key Events
```{r}
derive_events <- function(communication_tbl, round_tbl) {

  keywords <- c("civicloom", "harborcrest", "embargo",
                "anonymous", "saltwind", "bilateral", "#AlgorithmicEviction")

  bind_rows(
    
    communication_tbl %>%
      filter(
        public_channel | anon_channel | side_channel, 
        deal_terms | embargo_terms | crisis_terms,
        hour < ymd_hms("2046-06-05 18:00:00")
      ) %>%
      slice_min(hour, n = 1, with_ties = FALSE) %>%
      transmute(
        hour,
        label = paste0("BREACH:\n", agent_id, "\n(", channel, ")"),
        category = "breach"
      ), 
      
    communication_tbl %>%
      group_by(channel) %>%
      slice_min(hour, n = 1, with_ties = FALSE) %>%
      ungroup() %>%
      transmute(
        hour,
        label = paste0(channel, "\nfirst used"), # First appearance of each channel
        category = "channel"
      ),

    communication_tbl %>%
      mutate(
        kw = str_extract(str_to_lower(content),
                         paste(keywords, collapse = "|"))
      ) %>%
      filter(!is.na(kw)) %>%
      group_by(kw) %>%
      slice_min(hour, n = 1, with_ties = FALSE) %>%
      ungroup() %>%
      transmute(
        hour,
        label = paste0(str_to_title(kw), "\nfirst mentioned"), # First mention of keyword
        category = "keyword"
      ),

    round_tbl %>%
      filter(agents_unavailable != "") %>%
      transmute(
        hour,
        label = paste0("Absent:\n", agents_unavailable), # when agents were absent
        category = "oversight"
      ),

    round_tbl %>%
      mutate(vol_z = as.numeric(scale(n_comms))) %>%
      filter(vol_z > 1.5) %>%
      transmute(
        hour,
        label = paste0("Volume spike\n(z = ", round(vol_z, 1), ")"),
        category = "volume"
      )

  ) %>%
    arrange(hour) %>%
    distinct(hour, label, .keep_all = TRUE)
}

```




```{r}
events_auto <- derive_events(communication_tbl, round_tbl)

# Label the key events 
events_labelled <- events_auto %>%
  mutate(
    severity = case_when(
      category == "breach" ~ 6,
      category == "oversight" ~ 5,
      category == "keyword" ~ 4,
      category == "channel" ~ 3,
      category == "volume" ~ 2,
      TRUE ~ 1
    ),
    event_type = category  # or recode to friendlier names
  )

ggplot(events_labelled, aes(x = hour, y = severity, colour = event_type)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = label), size = 3, show.legend = FALSE) +
  scale_x_datetime(date_labels = "%d %b") +
  scale_y_continuous(breaks = 1:5, limits = c(1, 5.5)) +
  labs(title = "Event sequence and severity", x = NULL, y = "Severity",
       colour = "Event type") +
  theme_forensic()

```









