---
title: "Hands-on Exercise 1"
author: "Lei Yuhe"
date: "April 15, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Getting Started with ggplot2

Use ***p_load()*** function to install the collection of R packages under tidyverse, which will include the ggplot2, often used to create data-driven graphics based on **The Grammar of Graphics**. 

```{r}
pacman::p_load(tidyverse)
```

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```
We got 322 rows of data with 7 attributes. 

- The categorical attributes are: ID, CLASS, GENDER and RACE.
- The continuous attributes are: MATHS, ENGLISH and SCIENCE.

## Histogram of Maths Scores

Here, you can compare the difference between R Graphics, the core graphical function of Base R, and ggplot plotting a histogram. 

::: panel-tabset
### R Graphics

```{r}
hist(exam_data$MATHS)
```

### ggplot2

```{r}
ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 10, color = "black", fill = "lightgrey") +
  labs(title = "Histogram of exam_data$MATHS",
       x = "exam_data$MATHS",
       y = "Frequency") +
  theme_bw()
```
:::

As we can see, the difference is obvious. As quoted from [Hadley Wickham](http://varianceexplained.org/r/teach_ggplot2_to_beginners/#comment-1745406157): 

> The transferable skills from ggplot2 are not the idiosyncrasies of plotting syntax, but a powerful way of thinking about visualisation, as a way of mapping between variables and the visual properties of geometric objects that you can perceive.

## Grammar of Graphics
The ***ggplot2*** follows a layered grammer of graphics, the 7 layers can be described as follow: 

- **Data**: The dataset being plotted.
- **Aesthetics**: It take attributes of the data and use them to influence visual characteristics, such as position, colours, size, shape, or transparency.
- **Geometrics**: The visual elements used for our data, such as point, bar or line.
- **Facets**: Split the data into subsets to create multiple variations of the same graph (paneling, multiple plots).
- **Statistics**: Statiscal transformations that summarise data (e.g. mean, confidence intervals).
- **Coordinate**: It defines the plane on which data are mapped on the graphic.
- **Themes**: Modify all non-data components of a plot, such as main title, sub-title, y-aixs title, or legend background.


## Reference
- Prof. Kam Tin Seong [R for Visual Analytics](https://r4va.netlify.app/chap01#learning-outcome).
