---
title: "Hands-on Exercise 7a: Modelling, Visualising and Analysing Network Data with R"
author: "Lei Yuhe"
date: "May 14, 2026"
date-modified: "last-modified"
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    code-fold: show
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---


# Overview

This chapter covers how to **model, analyse and visualise network data** using R.

Objectives: 

- create graph object data frames and manipulate them with **dplyr**, **lubridate**, and **tidygraph**
- build network graph visualisations with **ggraph**
- compute network metrics (centrality, community) with **tidygraph**
- build advanced graph visualisations that incorporate these metrics
- build **interactive** network visualisations with **visNetwork**

Source data: an oil exploration/extraction company's internal email network
(the **GAStech** case data), consisting of an *edges* file (who emailed whom)
and a *nodes* file (employee attributes).

# Getting Started
## Installing and launching R packages

Packages needed for this chapter:

| Package | Purpose |
|---|---|
| `igraph` | Core network data structures & algorithms |
| `tidygraph` | Tidy (dplyr-style) API for graph manipulation |
| `ggraph` | ggplot2-style grammar for network graphs |
| `visNetwork` | Interactive network graphs (vis.js) |
| `tidyverse` | Data wrangling |
| `lubridate` | Date/time handling |
| `clock` | Additional date/time utilities |
| `graphlayouts` | Extra layout algorithms for ggraph |
| `concaveman` | Concave hulls (used with ggforce) |
| `ggforce` | Extra ggplot2 geoms, e.g. `geom_mark_hull()` |

```{r}
#| label: load-packages
pacman::p_load(igraph, tidygraph, ggraph,
                visNetwork, lubridate, clock,
                tidyverse, graphlayouts,
                concaveman, ggforce)
```

## Importing network data from files

```{r}
#| label: import-data
GAStech_nodes <- read_csv("data/GAStech_email_node.csv")
GAStech_edges <- read_csv("data/GAStech_email_edge-v2.csv")
```

## Reviewing the imported data

```{r}
#| label: glimpse-edges
glimpse(GAStech_edges)
```

::: {.callout-warning}
`SentDate` is imported as **character**, not **date**. This needs to be fixed
before any time-based analysis or visualisation can be done.
:::

## Wrangling time

```{r}
#| label: wrangle-time
GAStech_edges <- GAStech_edges %>%
  mutate(SendDate = dmy(SentDate)) %>%
  mutate(Weekday = wday(SentDate,
                         label = TRUE,
                         abbr = FALSE))
```

**Notes to self:**

- `dmy()` and `wday()` are both **lubridate** functions.
- `dmy()` parses a day-month-year character string into a proper `Date`.
- `wday()` extracts the weekday; `label = TRUE` returns names instead of
  numbers, and `abbr = FALSE` keeps full names (e.g. "Monday" not "Mon").
- The resulting `Weekday` column is an **ordinal** (ordered factor) field.

## Reviewing the revised date fields

```{r}
#| label: glimpse-edges-revised
glimpse(GAStech_edges)
```

## Wrangling attributes

Raw edges are individual email records, which is too granular for a clear network
plot. Aggregate by sender, receiver, and weekday to get an edge **weight**
(count of emails):

```{r}
#| label: aggregate-edges
GAStech_edges_aggregated <- GAStech_edges %>%
  filter(MainSubject == "Work related") %>%
  group_by(source, target, Weekday) %>%
  summarise(Weight = n()) %>%
  filter(source != target) %>%
  filter(Weight > 1) %>%
  ungroup()
```
**Notes to self:**

- Four **dplyr** verbs at work: `filter()`, `group_by()`, `summarise()`,
  `ungroup()`.
- `filter(source != target)` removes self-loops (emails to oneself).
- `filter(Weight > 1)` drops very weak (one-off) ties to reduce clutter.
- Output: **1,372 rows**, a much more usable edge list.

## Reviewing the revised edges file

```{r}
#| label: glimpse-aggregated
glimpse(GAStech_edges_aggregated)
```

# Creating Network Objects Using `tidygraph`

**tidygraph** gives a tidy API for graph manipulation. A network isn't
naturally "tidy," but it can be represented as **two** tidy tables (nodes +
edges), and tidygraph provides a clean way to move between them.

Recommended background reading:

- [Introducing tidygraph](https://www.data-imaginist.com/2017/introducing-tidygraph/)
- [tidygraph 1.1 — A tidy hope](https://www.data-imaginist.com/2018/tidygraph-1-1-a-tidy-hope/)

## The `tbl_graph` object

Two constructors:

- **`tbl_graph()`** build a `tbl_graph` directly from separate nodes and
  edges data frames.
- **`as_tbl_graph()`** convert an *existing* network object/data structure
  into a `tbl_graph`. Supported inputs include: node/edge data frame pairs,
  base R `data.frame`/`list`/`matrix`, `igraph` objects, `network` objects,
  `dendrogram`/`hclust`, `data.tree::Node`, `ape::phylo`/`evonet`, and
  Bioconductor `graphNEL`/`graphAM`/`graphBAM` objects.

## The `dplyr` verbs in `tidygraph`

- **`activate()`** switches which internal tibble (nodes or edges) is
  "active." Any dplyr verb applied to a `tbl_graph` acts on whichever tibble
  is currently active.
- While manipulating one tibble, you can still reach into the other:
  - `.N()` access the node data
  - `.E()` access the edge data
  - `.G()` access the whole `tbl_graph` object

## Using `tbl_graph()` to build the tidygraph data model

```{r}
#| label: build-tbl-graph
GAStech_graph <- tbl_graph(nodes = GAStech_nodes,
                            edges = GAStech_edges_aggregated,
                            directed = TRUE)
```

## Reviewing the output tidygraph's graph object

```{r}
#| label: print-graph
GAStech_graph
```

**Notes to self:**

- Printing a `tbl_graph` shows node count/edge count, and previews both
  tibbles.
- One tibble is always marked **(active)**, that's the one dplyr verbs will
  currently affect.

## Changing the active object

To reorder edges by descending weight, first `activate(edges)`: 

```{r}
#| label: activate-edges
GAStech_graph %>%
  activate(edges) %>%
  arrange(desc(Weight))
```

See the [`activate()` reference](https://tidygraph.data-imaginist.com/reference/activate.html)
for more.

# Plotting Static Network Graphs with `ggraph`

[**ggraph**](https://ggraph.data-imaginist.com/) extends **ggplot2**'s
grammar of graphics to networks. Three core building blocks:

- [nodes](https://cran.r-project.org/web/packages/ggraph/vignettes/Nodes.html)
- [edges](https://cran.r-project.org/web/packages/ggraph/vignettes/Edges.html)
- [layouts](https://cran.r-project.org/web/packages/ggraph/vignettes/Layouts.html)

## Plotting a basic network graph

```{r}
#| label: basic-graph
ggraph(GAStech_graph) +
  geom_edge_link() +
  geom_node_point()
```

**Notes to self:** `ggraph()` accepts either an `igraph` or a `tbl_graph`
object, plus a `layout` argument (defaults to `"nicely"` if unspecified).

## Changing the default network graph theme

```{r}
#| label: theme-graph
g <- ggraph(GAStech_graph) +
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

`theme_graph()` strips axes/grid/border and swaps in a network-friendly
default font (Arial Narrow, overridable). Can be applied per-plot with
`theme_graph()` or globally with `set_graph_style()`.

## Changing the coloring of the plot

```{r}
#| label: theme-graph-color
g <- ggraph(GAStech_graph) +
  geom_edge_link(aes(colour = 'grey50')) +
  geom_node_point(aes(colour = 'grey40'))

g + theme_graph(background = 'grey10',
                text_colour = 'white')
```

## Working with `ggraph`'s layouts

Supported standard layouts: `star`, `circle`, `nicely` (default), `dh`,
`gem`, `graphopt`, `grid`, `mds`, `sphere`, `randomly`, `fr`, `kk`, `drl`,
`lgl`.

## Fruchterman and Reingold layout

```{r}
#| label: fr-layout
g <- ggraph(GAStech_graph,
            layout = "fr") +
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

`layout = "fr"` selects the Fruchterman–Reingold force-directed layout.

## Modifying network nodes

Colour nodes by `Department`:

```{r}
#| label: color-nodes-department
g <- ggraph(GAStech_graph,
            layout = "nicely") +
  geom_edge_link(aes()) +
  geom_node_point(aes(colour = Department,
                      size = 3))

g + theme_graph()
```

`geom_node_point()` is the network analogue of `geom_point()` — supports
colour, size, shape aesthetics just like ggplot2.

## Modifying edges

Map edge thickness to `Weight`:

```{r}
#| label: edge-width
g <- ggraph(GAStech_graph,
            layout = "nicely") +
  geom_edge_link(aes(width = Weight),
                 alpha = 0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department),
                  size = 3)

g + theme_graph()
```

`width` maps line thickness proportionally to `Weight`; `alpha` adds
transparency to reduce over-plotting.


# Creating Facet Graphs

Faceting spreads nodes/edges across panels by attribute value, reducing
edge over-plotting. Three ggraph faceting functions:

- **`facet_nodes()`** - an edge is drawn in a panel only if *both* its
  terminal nodes are present in that panel.
- **`facet_edges()`** - nodes are drawn in *every* panel, regardless of the
  faceting variable.
- **`facet_graph()`** - facet on two variables simultaneously.

## Working with `facet_edges()`

```{r}
#| label: facet-edges-1
set_graph_style()

g <- ggraph(GAStech_graph,
            layout = "nicely") +
  geom_edge_link(aes(width = Weight),
                 alpha = 0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department),
                  size = 2)

g + facet_edges(~Weekday)
```

## Repositioning the legend

```{r}
#| label: facet-edges-legend
#| fig-width: 10
set_graph_style()

g <- ggraph(GAStech_graph,
            layout = "nicely") +
  geom_edge_link(aes(width = Weight),
                 alpha = 0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department),
                  size = 2) +
  theme(legend.position = 'bottom')

g + facet_edges(~Weekday)
```

## A framed facet graph

```{r}
#| label: facet-edges-framed
#| fig-width: 10
set_graph_style()

g <- ggraph(GAStech_graph,
            layout = "nicely") +
  geom_edge_link(aes(width = Weight),
                 alpha = 0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department),
                  size = 2)

g + facet_edges(~Weekday) +
  th_foreground(foreground = "grey80",
                border = TRUE) +
  theme(legend.position = 'bottom')
```

## Working with `facet_nodes()`

```{r}
#| label: facet-nodes
#| fig-width: 10
set_graph_style()

g <- ggraph(GAStech_graph,
            layout = "nicely") +
  geom_edge_link(aes(width = Weight),
                 alpha = 0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department),
                  size = 2)

g + facet_nodes(~Department) +
  th_foreground(foreground = "grey80",
                border = TRUE) +
  theme(legend.position = 'bottom')
```

# Network Metrics Analysis

## Computing centrality indices

Centrality measures describe how "important" an actor is within a network.
Four classic measures: **degree, betweenness, closeness, eigenvector**.

::: {.callout-note}
For the underlying theory/maths, see *Chapter 7: Actor Prominence* in
**A User's Guide to Network Analysis in R**.
:::

```{r}
#| label: betweenness-centrality
#| fig-height: 6
g <- GAStech_graph %>%
  mutate(betweenness_centrality = centrality_betweenness()) %>%
  ggraph(layout = "fr") +
  geom_edge_link(aes(width = Weight),
                 alpha = 0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department,
                       size = betweenness_centrality))
g + theme_graph()
```

`mutate()` (dplyr) + `centrality_betweenness()` (tidygraph) computes and
stores betweenness centrality as a new node attribute.

## Visualising network metrics

From **ggraph v2.0** onward, tidygraph algorithms (like centrality
functions) can be called **directly inside** a ggraph `aes()` mapping, no
need to precompute and store them first:

```{r}
#| label: centrality-inline
#| fig-height: 6
g <- GAStech_graph %>%
  ggraph(layout = "fr") +
  geom_edge_link(aes(width = Weight),
                 alpha = 0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department,
                       size = centrality_betweenness()))
g + theme_graph()
```

## Visualising community

tidygraph exposes many igraph community-detection algorithms:

- Edge-betweenness - `group_edge_betweenness()`
- Leading eigenvector — `group_leading_eigen()`
- Fast-greedy — `group_fast_greedy()`
- Louvain — `group_louvain()`
- Walktrap — `group_walktrap()`
- Label propagation — `group_label_prop()`
- InfoMAP — `group_infomap()`
- Spinglass — `group_spinglass()`
- Optimal — `group_optimal()`

Some account for edge direction/weight, some don't — see the
[tidygraph community reference](https://tidygraph.data-imaginist.com/reference/group_graph.html).

```{r}
#| label: community-edge-betweenness
#| fig-height: 6
g <- GAStech_graph %>%
  mutate(community = as.factor(
    group_edge_betweenness(
      weights = Weight,
      directed = TRUE))) %>%
  ggraph(layout = "fr") +
  geom_edge_link(
    aes(width = Weight),
    alpha = 0.2) +
  scale_edge_width(
    range = c(0.1, 5)) +
  geom_node_point(
    aes(colour = community))

g + theme_graph()
```

### Adding hulls around communities

Uses `geom_mark_hull()` from **ggforce** (requires **ggforce** +
**concaveman**) to visually enclose each detected community:

```{r}
#| label: community-hulls
#| fig-height: 6
g <- GAStech_graph %>%
  activate(nodes) %>%
  mutate(community = as.factor(
    group_optimal(weights = Weight)),
         betweenness_measure = centrality_betweenness()) %>%
  ggraph(layout = "fr") +
  geom_mark_hull(
    aes(x, y,
        group = community,
        fill = community),
    alpha = 0.2,
    expand = unit(0.3, "cm"),   # how far the hull extends beyond points
    radius = unit(0.3, "cm")   # corner smoothness
  ) +
  geom_edge_link(aes(width = Weight),
                 alpha = 0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(fill = Department,
                       size = betweenness_measure),
                  color = "black",
                  shape = 21)

g + theme_graph()
```

# Building Interactive Network Graphs with `visNetwork`

[**visNetwork**](http://datastorm-open.github.io/visNetwork/) wraps the
[vis.js](http://visjs.org) JavaScript library for interactive network
visualisation.

Key requirements:

- The nodes list must include an **`id`** column.
- The edges list must have **`from`** and **`to`** columns.
- Node labels are drawn from the nodes list's **`label`** column.
- Interactive features: drag nodes (auto re-spacing), zoom, and pan.

## Data preparation

```{r}
#| label: visnetwork-data-prep
GAStech_edges_aggregated <- GAStech_edges %>%
  left_join(GAStech_nodes, by = c("sourceLabel" = "label")) %>%
  rename(from = id) %>%
  left_join(GAStech_nodes, by = c("targetLabel" = "label")) %>%
  rename(to = id) %>%
  filter(MainSubject == "Work related") %>%
  group_by(from, to) %>%
  summarise(weight = n()) %>%
  filter(from != to) %>%
  filter(weight > 1) %>%
  ungroup()
```

## Plotting the first interactive network graph

```{r}
#| label: visnetwork-basic
#| eval: false
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated)
```

## Working with layout

```{r}
#| label: visnetwork-fr-layout
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr")
```

See [`visIgraphLayout` docs](http://datastorm-open.github.io/visNetwork/igraph.html)
for available layout options.

## Working with visual attributes — Nodes

`visNetwork()` colours nodes by a field named **`group`**, so rename
`Department` to `group`:

```{r}
#| label: visnetwork-group-rename
GAStech_nodes <- GAStech_nodes %>%
  rename(group = Department)
```

```{r}
#| label: visnetwork-colored
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

## Working with visual attributes — Edges

`visEdges()` controls edge appearance:

- `arrows` — where to place arrowheads (e.g. `"to"`)
- `smooth` — draw edges as smooth/curved lines

```{r}
#| label: visnetwork-edges
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visEdges(arrows = "to",
           smooth = list(enabled = TRUE,
                         type = "curvedCW")) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

See [visEdges options](http://datastorm-open.github.io/visNetwork/edges.html)
for more.

## Interactivity

`visOptions()` adds interactive behaviours:

- `highlightNearest = TRUE` — highlight a node's neighbours on click
- `nodesIdSelection = TRUE` — adds an HTML dropdown to jump to a node by id

```{r}
#| label: visnetwork-interactive
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visOptions(highlightNearest = TRUE,
             nodesIdSelection = TRUE) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

See [visOptions](http://datastorm-open.github.io/visNetwork/options.html)
for the full argument list.

# Key Takeaways

1. **Two-table model**: every network = one nodes table + one edges table;
   `tidygraph::tbl_graph()` binds them into one object.
2. **`activate()`** decides which table (nodes/edges) your dplyr verbs act
   on — a very easy thing to forget mid-pipeline.
3. **ggraph = ggplot2 for networks**: same `aes()`/geom logic, with
   network-specific geoms (`geom_node_point()`, `geom_edge_link()`) and
   layouts (`fr`, `kk`, `nicely`, etc.).
4. **Faceting** (`facet_edges()`, `facet_nodes()`) is a powerful way to
   reduce over-plotting by splitting the network across a categorical or
   temporal variable.
5. **Centrality & community detection** can be computed inline inside a
   `ggraph()` call (v2.0+) — no separate `mutate()` step required.
6. **visNetwork** trades static ggraph plots for browser-based
   interactivity (pan/zoom/drag/highlight), useful for exploratory
   analysis or dashboards.

# Reference

Course site: [R for Visual Analytics — Chapter 27](https://r4va.netlify.app/chap27)


















