---
title: "1. Visual Analytics for Insurance Portfolio Data Discovery"
author: "Lei Yuhe"
date: "May 17, 2026"
date-modified: "last-modified"
format: html
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---



# Data Understanding & Preparation

## Loading Libraries

The following R libraries are used:

| Library | Purpose |
|---|---|
| `tidyverse` | Data wrangling (`dplyr`, `tidyr`) and base visualisation (`ggplot2`) |
| `ggdist` | Visualising distributions and uncertainty (half-eye plots, dot plots) |
| `ggridges` | Density ridge plots for comparing distributions across groups |
| `ggstatsplot` | Statistical visualisations with built-in hypothesis testing |
| `ggmosaic` | Mosaic plots for categorical variable relationships |
| `patchwork` | Composing multiple ggplot objects into one figure |
| `scales` | Formatting axis labels (percentages, currency) |
| `knitr` | Rendering tables neatly within the report |
| `corrplot` | Correlation matrix visualisation |
| `forcats` | Factor reordering for cleaner categorical plots |

```{r}
#| label: load-libraries
#| code-fold: true
#| code-summary: "Show Code"

pacman::p_load(
  tidyverse,
  ggdist,
  ggridges,
  ggstatsplot,
  ggmosaic,
  patchwork,
  scales,
  knitr,
  corrplot,
  forcats,
  ggrepel,
  ggthemes
)
```

## Loading the Data

The dataset is from [Mendeley Data](https://data.mendeley.com/datasets/sw4jmdb2sm/1). It contains policy-level records from a Spanish non-life motor vehicle insurer.

```{r}
#| label: load-data

insurance <- read_delim("data/Dataset of motor insurance portfolio.csv", delim = ";")

# Preview structure
glimpse(insurance)
```

## Data Dictionary

Based on the dataset, the key variables are:

| Variable | Type | Description |
|---|---|---|
| `insured_id` | Identifier | Unique policyholder identifier |
| `year` | Numerical | Policy year (2022–2024) |
| `policy_type` | Categorical | Coverage type (COMP_E, COMP_N, CC, TPG, etc.) |
| `policy_status` | Categorical | Policy status (A = Active, C = Cancelled) |
| `business_type` | Categorical | New business (NB) or renewal (P) |
| `payment_frequency` | Categorical | Payment frequency (A = Annual, S = Semi-annual, etc.) |
| `bonus_score` | Categorical | Bonus-malus score reflecting claim history |
| `driver_age` | Numerical | Age of the main driver (years) |
| `vehicle_age` | Numerical | Age of the insured vehicle (years) |
| `age_driving_licence` | Numerical | Years since driving licence obtained |
| `fuel_type` | Categorical | Fuel type (G = Gasoline, D = Diesel, etc.) |
| `vehicle_value` | Numerical | Market value of insured vehicle (EUR) |
| `seats` | Numerical | Number of seats in the vehicle |
| `power_to_weight_ratio` | Numerical | Engine power-to-weight ratio |
| `vehicle_brand` | Categorical | Vehicle manufacturer |
| `municipality_type` | Categorical | Urban (I) or rural (C) classification |
| `circulation_area` | Categorical | Urban (U) or rural (R) circulation area |
| `total_premium` | Numerical | Total gross written premium (EUR) |
| `total_claims` | Numerical | Total number of claims across all coverage types |
| `total_incurred` | Numerical | Total incurred claim cost (EUR) |
| `total_exposure` | Numerical | Policy exposure (fraction of year in force) |


## Deriving Key Variables

Before visualisation, two target variables were derived:

- **Loss Ratio** = `claimed amount / premium`, which measures profitability. A ratio > 1 indicates the insurer is paying out more than it earns.
- **Claim Frequency** = `number of claims / exposure`, which measures volatility by normalising claims to exposure.

```{r}
#| label: feature-engineering
#| code-fold: true
#| code-summary: "Show Code"

insurance <- insurance %>%
  mutate(
    # Core profitability metric
    loss_ratio = total_incurred / total_premium,

    # Core volatility metric
    claim_frequency = total_claims / total_exposure,

    # Whether the policy had at least one claim
    had_claim = if_else(total_claims > 0, "With Claim", "No Claim"),

    # Driver age bands
    driver_age_group = cut(
      driver_age,
      breaks = c(17, 25, 35, 45, 55, 65, Inf),
      labels = c("18-25", "26-35", "36-45", "46-55", "56-65", "65+"),
      right  = TRUE
    ),

    # Vehicle age bands
    vehicle_age_group = cut(
      vehicle_age,
      breaks = c(-1, 2, 5, 10, 15, Inf),
      labels = c("0-2 yrs", "3-5 yrs", "6-10 yrs", "11-15 yrs", "15+ yrs"),
      right  = TRUE
    ),

    # Simplify circulation area labels
    area_label = case_when(
      circulation_area == "U" ~ "Urban",
      circulation_area == "R" ~ "Rural",
      TRUE ~ circulation_area
    ),

    # Simplify fuel type labels
    fuel_label = case_when(
      fuel_type == "G" ~ "Gasoline",
      fuel_type == "D" ~ "Diesel",
      fuel_type == "E" ~ "Electric",
      fuel_type == "H" ~ "Hybrid",
      TRUE ~ fuel_type
    ),

    # Simplify business type labels
    business_label = case_when(
      business_type == "NB" ~ "New Business",
      business_type == "P"  ~ "Renewal",
      TRUE ~ business_type
    )
  )
```

## Data Quality Check

```{r}
#| label: data-quality
#| code-fold: true
#| code-summary: "Show Code"

# Missing values
missing_summary <- insurance %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(),
               names_to  = "Variable",
               values_to = "Missing_Count") %>%
  filter(Missing_Count > 0) %>%
  arrange(desc(Missing_Count))

if (nrow(missing_summary) == 0) {
  cat("No missing values detected.\n")
} else {
  kable(missing_summary, caption = "Variables with Missing Values")
}

# Summary of core numerical variables
core_nums <- insurance %>%
  select(driver_age, vehicle_age, vehicle_value, total_premium,
         total_incurred, total_claims, total_exposure,
         loss_ratio, claim_frequency)

kable(summary(core_nums), caption = "Summary Statistics — Core Numerical Variables")
```

## Outlier Treatment

```{r}
#| label: outlier-handling
#| code-fold: true
#| code-summary: "Show Code"

lr_upper <- quantile(insurance$loss_ratio,      0.99, na.rm = TRUE)
cf_upper <- quantile(insurance$claim_frequency, 0.99, na.rm = TRUE)

cat("Loss Ratio 99th percentile:      ", round(lr_upper, 4), "\n")
cat("Claim Frequency 99th percentile: ", round(cf_upper, 4), "\n")

insurance <- insurance %>%
  mutate(
    is_extreme_lr = loss_ratio      > lr_upper,
    is_extreme_cf = claim_frequency > cf_upper
  )

cat("Policies with extreme LR (> 99th pct):", sum(insurance$is_extreme_lr, na.rm = TRUE), "\n")
cat("Policies with extreme CF (> 99th pct):", sum(insurance$is_extreme_cf, na.rm = TRUE), "\n")

# Clean dataset excludes extremes for primary visualisation
ins_clean <- insurance %>%
  filter(!is_extreme_lr, !is_extreme_cf)
```

---


# Univariate Visual Analytics

## Setting a Common Theme

```{r}
#| label: custom-theme
#| code-fold: true
#| code-summary: "Show Code"

theme_ins <- theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(size = 10, colour = "grey40"),
    axis.title       = element_text(size = 10),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )

# Colour palettes
policy_colours <- c(
  "COMP_E" = "#4E79A7", "COMP_N" = "#F28E2B",
  "CC"     = "#59A14F", "TPG"    = "#E15759",
  "TP"     = "#76B7B2"
)

fuel_colours <- c(
  "Gasoline" = "#4E79A7", "Diesel" = "#F28E2B",
  "Electric" = "#59A14F", "Hybrid" = "#E15759"
)
```



## Numerical Variables

### Total Premium Distribution

```{r}
#| label: vis1-premium
#| fig-cap: "Distribution of Total Gross Written Premium"
#| code-fold: true
#| code-summary: "Show Code"

p_prem_hist <- ggplot(ins_clean, aes(x = total_premium)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 60, fill = "#4E79A7",
                 colour = "white", alpha = 0.85) +
  geom_density(colour = "#E15759", linewidth = 1) +
  scale_x_continuous(labels = label_dollar(prefix = "€")) +
  labs(title    = "Distribution of Total Premium",
       subtitle = "Right-skewed; most policies priced under €1,000",
       x = "Total Premium (EUR)", y = "Density") +
  theme_ins

p_prem_box <- ggplot(ins_clean, aes(y = total_premium, x = "")) +
  geom_violin(fill = "#4E79A7", alpha = 0.5) +
  geom_boxplot(width = 0.12, fill = "white", outlier.shape = NA) +
  scale_y_continuous(labels = label_dollar(prefix = "€")) +
  coord_flip() +
  labs(x = NULL, y = "Total Premium (EUR)", title = "Spread") +
  theme_ins

p_prem_hist / p_prem_box + plot_layout(heights = c(3, 1))
```


### Total Incurred Claim Cost (Claimants Only)

```{r}
#| label: vis2-incurred
#| fig-cap: "Distribution of Total Incurred Claim Cost"
#| code-fold: true
#| code-summary: "Show Code"

ins_claimants <- ins_clean %>% filter(total_claims > 0)

p_inc_raw <- ggplot(ins_claimants, aes(x = total_incurred)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 60, fill = "#E15759",
                 colour = "white", alpha = 0.85) +
  geom_density(colour = "#4E79A7", linewidth = 1) +
  scale_x_continuous(labels = label_dollar(prefix = "€")) +
  labs(title    = "Raw Total Incurred (Claimants Only)",
       subtitle = "Extreme right skew; most claims are small",
       x = "Total Incurred (EUR)", y = "Density") +
  theme_ins

p_inc_log <- ggplot(ins_claimants, aes(x = log1p(total_incurred))) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 50, fill = "#F28E2B",
                 colour = "white", alpha = 0.85) +
  geom_density(colour = "#4E79A7", linewidth = 1) +
  labs(title    = "Log-Transformed Total Incurred",
       subtitle = "Log scale reveals approximate log-normal structure",
       x = "log(Total Incurred + 1)", y = "Density") +
  theme_ins

p_inc_raw + p_inc_log
```

### Loss Ratio Distribution

```{r}
#| label: vis3-loss-ratio
#| fig-cap: "Policy-Level Loss Ratio Distribution"
#| code-fold: true
#| code-summary: "Show Code"

ggplot(ins_clean, aes(x = loss_ratio)) +
  geom_histogram(
    aes(y = after_stat(density), fill = after_stat(x) > 1),
    bins = 80, colour = "white", alpha = 0.85
  ) +
  scale_fill_manual(
    values = c("FALSE" = "#59A14F", "TRUE" = "#E15759"),
    labels = c("Profitable (LR < 1)", "Unprofitable (LR ≥ 1)"),
    name   = NULL
  ) +
  geom_vline(xintercept = 1, linetype = "dashed",
             colour = "black", linewidth = 0.9) +
  geom_density(colour = "grey30", linewidth = 0.8) +
  annotate("text", x = 1.05, y = Inf, vjust = 2, hjust = 0,
           label = "Break-even (LR = 1)", size = 3.5, colour = "grey30") +
  scale_x_continuous(labels = percent_format()) +
  labs(title    = "Distribution of Loss Ratio Across Policies",
       subtitle = "Policies with LR > 1 represent net losses to the insurer",
       x = "Loss Ratio", y = "Density") +
  theme_ins
```


### Driver Age and Vehicle Age Distributions

```{r}
#| label: vis4-age
#| fig-cap: "Driver Age and Vehicle Age Distributions"
#| code-fold: true
#| code-summary: "Show Code"

p_driver <- ggplot(ins_clean, aes(x = driver_age)) +
  geom_histogram(bins = 40, fill = "#4E79A7",
                 colour = "white", alpha = 0.85) +
  labs(title    = "Driver Age Distribution",
       subtitle = "Bell-shaped; centred around middle-aged drivers",
       x = "Driver Age (years)", y = "Count") +
  theme_ins

p_vehicle <- ggplot(ins_clean, aes(x = vehicle_age)) +
  geom_histogram(bins = 30, fill = "#F28E2B",
                 colour = "white", alpha = 0.85) +
  labs(title    = "Vehicle Age Distribution",
       subtitle = "Right-skewed; portfolio skews toward newer vehicles",
       x = "Vehicle Age (years)", y = "Count") +
  theme_ins

p_driver + p_vehicle
```

### Vehicle Value Distribution

```{r}
#| label: vis5-vehicle-value
#| fig-cap: "Distribution of Vehicle Value"
#| code-fold: true
#| code-summary: "Show Code"

vv_upper <- quantile(ins_clean$vehicle_value, 0.99, na.rm = TRUE)

ggplot(ins_clean %>% filter(vehicle_value <= vv_upper),
       aes(x = vehicle_value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = "#76B7B2", colour = "white", alpha = 0.85) +
  geom_density(colour = "#E15759", linewidth = 1) +
  scale_x_continuous(labels = label_dollar(prefix = "€")) +
  labs(title    = "Vehicle Value Distribution (up to 99th percentile)",
       subtitle = "Wide spread; luxury brands (Porsche, Mercedes) create a long tail",
       x = "Vehicle Value (EUR)", y = "Density") +
  theme_ins
```


## Categorical Variables

### Policy Type Breakdown

```{r}
#| label: vis6-policy-type
#| fig-cap: "Policy Count and Premium Contribution by Policy Type"
#| code-fold: true
#| code-summary: "Show Code"

p_type_count <- ins_clean %>%
  count(policy_type) %>%
  mutate(policy_type = fct_reorder(policy_type, n)) %>%
  ggplot(aes(x = n, y = policy_type, fill = policy_type)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = comma(n)), hjust = -0.1, size = 3.5) +
  scale_fill_manual(values = policy_colours) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Policy Count by Policy Type",
       x = "Number of Policies", y = NULL) +
  theme_ins

p_type_prem <- ins_clean %>%
  group_by(policy_type) %>%
  summarise(total = sum(total_premium)) %>%
  mutate(
    policy_type = fct_reorder(policy_type, total),
    pct = total / sum(total)
  ) %>%
  ggplot(aes(x = total, y = policy_type, fill = policy_type)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = percent(pct, accuracy = 0.1)), hjust = -0.1, size = 3.5) +
  scale_fill_manual(values = policy_colours) +
  scale_x_continuous(labels = label_dollar(prefix = "€", suffix = "M", scale = 1e-6),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Total Premium by Policy Type",
       x = "Total Premium (EUR Millions)", y = NULL) +
  theme_ins

p_type_count + p_type_prem
```

### Business Type and Payment Frequency

```{r}
#| label: vis7-business-payment
#| fig-cap: "New Business vs Renewal and Payment Frequency Breakdown"
#| code-fold: true
#| code-summary: "Show Code"

p_biz <- ins_clean %>%
  count(business_label) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(x = "", y = pct, fill = business_label)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = paste0(business_label, "\n", percent(pct, accuracy = 0.1))),
            position = position_stack(vjust = 0.5), size = 3.5) +
  scale_fill_manual(values = c("New Business" = "#4E79A7", "Renewal" = "#59A14F")) +
  coord_flip() +
  labs(title = "New Business vs Renewal Split", x = NULL, y = NULL) +
  theme_ins + theme(legend.position = "none",
                    axis.text = element_blank(),
                    axis.ticks = element_blank())

p_pay <- ins_clean %>%
  count(payment_frequency) %>%
  mutate(
    payment_frequency = fct_reorder(payment_frequency, n),
    pct = n / sum(n)
  ) %>%
  ggplot(aes(x = pct, y = payment_frequency, fill = payment_frequency)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = percent(pct, accuracy = 0.1)), hjust = -0.1, size = 3.5) +
  scale_x_continuous(labels = percent_format(),
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Blues") +
  labs(title = "Payment Frequency Breakdown",
       x = "Share of Portfolio", y = "Payment Frequency") +
  theme_ins

p_biz + p_pay
```


### Fuel Type and Circulation Area

```{r}
#| label: vis8-fuel-area
#| fig-cap: "Fuel Type and Circulation Area Composition"
#| code-fold: true
#| code-summary: "Show Code"

p_fuel <- ins_clean %>%
  count(fuel_label) %>%
  mutate(
    fuel_label = fct_reorder(fuel_label, n),
    pct = n / sum(n)
  ) %>%
  ggplot(aes(x = pct, y = fuel_label, fill = fuel_label)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = percent(pct, accuracy = 0.1)), hjust = -0.1, size = 3.5) +
  scale_fill_manual(values = fuel_colours) +
  scale_x_continuous(labels = percent_format(),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Fuel Type Composition",
       x = "Share of Portfolio", y = NULL) +
  theme_ins

p_area <- ins_clean %>%
  count(area_label) %>%
  mutate(
    area_label = fct_reorder(area_label, n),
    pct = n / sum(n)
  ) %>%
  ggplot(aes(x = pct, y = area_label, fill = area_label)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = percent(pct, accuracy = 0.1)), hjust = -0.1, size = 3.5) +
  scale_fill_manual(values = c("Urban" = "#4E79A7", "Rural" = "#59A14F")) +
  scale_x_continuous(labels = percent_format(),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Circulation Area Composition",
       x = "Share of Portfolio", y = NULL) +
  theme_ins

p_fuel + p_area
```

# Bivariate Visual Analytics

## Profitability Analysis (Loss Ratio)

### Loss Ratio by Policy Type using Raincloud Plot

```{r}
#| label: vis9-lr-policy-type
#| fig-cap: "Loss Ratio Distribution by Policy Type (Raincloud Plot)"
#| code-fold: true
#| code-summary: "Show Code"

ins_clean %>%
  filter(!is.na(policy_type)) %>%
  mutate(policy_type = fct_reorder(policy_type, loss_ratio, .fun = median)) %>%
  ggplot(aes(x = policy_type, y = loss_ratio, fill = policy_type)) +
  stat_halfeye(
    adjust = 0.5, justification = -0.2,
    .width = 0, point_colour = NA, alpha = 0.7
  ) +
  geom_boxplot(width = 0.15, outlier.shape = NA, colour = "grey30") +
  stat_dots(side = "left", justification = 1.1,
            binwidth = 0.008, dotsize = 1.5) +
  geom_hline(yintercept = 1, linetype = "dashed",
             colour = "red", linewidth = 0.7) +
  scale_fill_manual(values = policy_colours) +
  scale_y_continuous(labels = percent_format()) +
  coord_flip() +
  labs(title    = "Loss Ratio Distribution by Policy Type",
       subtitle = "Red dashed line = break-even (LR = 1); raincloud shows full distribution",
       x = NULL, y = "Loss Ratio") +
  theme_ins +
  theme(legend.position = "none")
```


### Statistical Test: Loss Ratio by Policy Type

```{r}
#| label: vis10-anova-lr
#| fig-cap: "One-Way ANOVA of Loss Ratio by Policy Type"
#| code-fold: true
#| code-summary: "Show Code"
#| fig-height: 10

ggbetweenstats(
  data  = ins_clean,
  x     = policy_type,
  y     = loss_ratio,
  type  = "nonparametric",
  p.adjust.method = "bonferroni",
  title = "Loss Ratio by Policy Type: Kruskal-Wallis Test",
  xlab  = "Policy Type",
  ylab  = "Loss Ratio",
  messages = FALSE
)
```

### Loss Ratio by Driver Age Group using Raincloud Plot

```{r}
#| label: vis11-lr-driver-age
#| fig-cap: "Loss Ratio Distribution by Driver Age Group"
#| code-fold: true
#| code-summary: "Show Code"

ins_clean %>%
  filter(!is.na(driver_age_group)) %>%
  ggplot(aes(x = driver_age_group, y = loss_ratio, fill = driver_age_group)) +
  stat_halfeye(adjust = 0.5, justification = -0.2,
               .width = 0, point_colour = NA, alpha = 0.7) +
  geom_boxplot(width = 0.15, outlier.shape = NA, colour = "grey30") +
  geom_hline(yintercept = 1, linetype = "dashed",
             colour = "red", linewidth = 0.7) +
  scale_fill_brewer(palette = "Spectral") +
  scale_y_continuous(labels = percent_format()) +
  coord_flip() +
  labs(title    = "Loss Ratio by Driver Age Group",
       subtitle = "Younger and older drivers typically exhibit higher, more volatile loss ratios",
       x = "Driver Age Group", y = "Loss Ratio") +
  theme_ins +
  theme(legend.position = "none")
```

### Loss Ratio by Circulation Area and Fuel Type using Heatmap

```{r}
#| label: vis12-heatmap-lr
#| fig-cap: "Mean Loss Ratio: Circulation Area vs Fuel Type"
#| code-fold: true
#| code-summary: "Show Code"

heatmap_data <- ins_clean %>%
  group_by(area_label, fuel_label) %>%
  summarise(
    mean_lr = mean(loss_ratio, na.rm = TRUE),
    n       = n(),
    .groups = "drop"
  ) %>%
  filter(!is.na(area_label), !is.na(fuel_label))

ggplot(heatmap_data,
       aes(x = fuel_label, y = area_label, fill = mean_lr)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(
    aes(label = paste0(percent(mean_lr, accuracy = 0.1),
                       "\n(n=", comma(n), ")")),
    size = 3.2, fontface = "bold"
  ) +
  scale_fill_gradient2(
    low = "#59A14F", mid = "#FFFF99", high = "#E15759",
    midpoint = 1,
    labels   = percent_format(),
    name     = "Mean Loss Ratio"
  ) +
  labs(title    = "Mean Loss Ratio: Circulation Area × Fuel Type",
       subtitle = "Green = profitable (< 100%); Red = unprofitable (> 100%)",
       x = "Fuel Type", y = "Circulation Area") +
  theme_ins
```

### Loss Ratio by Business Type

```{r}
#| label: vis13-lr-business
#| fig-cap: "Loss Ratio by Business Type (New vs Renewal)"
#| code-fold: true
#| code-summary: "Show Code"


ggbetweenstats(
  data  = ins_clean %>% filter(!is.na(business_label)),
  x     = business_label,
  y     = loss_ratio,
  type  = "nonparametric",
  title = "Loss Ratio: New Business vs Renewal",
  xlab  = "Business Type",
  ylab  = "Loss Ratio",
  messages = FALSE
)
```


## Volatility Analysis (Claim Frequency)

### Claim Frequency by Fuel Type

```{r}
#| label: vis14-cf-fuel
#| fig-cap: "Claim Frequency by Fuel Type"
#| code-fold: true
#| code-summary: "Show Code"


ins_clean %>%
  filter(!is.na(fuel_label)) %>%
  mutate(fuel_label = fct_reorder(fuel_label, claim_frequency, .fun = median)) %>%
  ggplot(aes(x = fuel_label, y = claim_frequency, fill = fuel_label)) +
  stat_halfeye(adjust = 0.5, justification = -0.2,
               .width = 0, point_colour = NA, alpha = 0.7) +
  geom_boxplot(width = 0.15, outlier.shape = NA, colour = "grey30") +
  scale_fill_manual(values = fuel_colours) +
  coord_flip() +
  labs(title    = "Claim Frequency Distribution by Fuel Type",
       subtitle = "Ordered by median claim frequency",
       x = "Fuel Type", y = "Claim Frequency (Claims per Exposure Year)") +
  theme_ins +
  theme(legend.position = "none")
```


### Vehicle Age vs. Loss Ratio: Risk Profile Bubble Chart

```{r}
#| label: vis15-bubble
#| fig-cap: "Risk Profile by Vehicle Age Group"
#| code-fold: true
#| code-summary: "Show Code"

risk_summary <- ins_clean %>%
  filter(!is.na(vehicle_age_group)) %>%
  group_by(vehicle_age_group) %>%
  summarise(
    n_policies      = n(),
    avg_cf          = mean(claim_frequency, na.rm = TRUE),
    avg_lr          = mean(loss_ratio, na.rm = TRUE),
    pct_with_claim  = mean(total_claims > 0, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(risk_summary,
       aes(x = avg_cf, y = avg_lr,
           size = n_policies, colour = vehicle_age_group)) +
  geom_point(alpha = 0.85) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
  geom_text_repel(aes(label = vehicle_age_group), size = 3.5, show.legend = FALSE) +
  scale_size_continuous(name = "Policy Count", range = c(4, 16)) +
  scale_colour_brewer(palette = "Set2", guide = "none") +
  scale_x_continuous(labels = number_format(accuracy = 0.01)) +
  scale_y_continuous(labels = percent_format()) +
  labs(title    = "Risk Profile by Vehicle Age Group",
       subtitle = "Top-right = high frequency & high loss ratio (worst segment); dashed = break-even",
       x = "Average Claim Frequency",
       y = "Average Loss Ratio") +
  theme_ins
```


### Loss Ratio by Bonus Score: Ridgeline Plot

```{r}
#| label: vis16-ridgeline-bonus
#| fig-cap: "Loss Ratio Distribution by Bonus-Malus Score"
#| code-fold: true
#| code-summary: "Show Code"

bonus_median <- ins_clean %>%
  filter(!is.na(bonus_score)) %>%
  group_by(bonus_score) %>%
  summarise(med_lr = median(loss_ratio, na.rm = TRUE)) %>%
  arrange(desc(med_lr))

ins_clean %>%
  filter(!is.na(bonus_score)) %>%
  mutate(bonus_score = factor(bonus_score,
                              levels = bonus_median$bonus_score)) %>%
  ggplot(aes(x = loss_ratio, y = bonus_score,
             fill = stat(quantile))) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE, quantiles = 4, quantile_lines = TRUE
  ) +
  geom_vline(xintercept = 1, linetype = "dashed",
             colour = "red", linewidth = 0.7) +
  scale_fill_manual(
    name   = "Quartile",
    values = c("#d4f1c5", "#c0c0c0", "#c0c0c0", "#f7b6b6")
  ) +
  scale_x_continuous(labels = percent_format()) +
  labs(title    = "Loss Ratio by Bonus-Malus Score",
       subtitle = "Ordered by median LR; bonus score reflects prior claim history",
       x = "Loss Ratio", y = "Bonus Score") +
  theme_ins
```

### Correlation Matrix of Numerical Variables

```{r}
#| label: vis17-correlation
#| fig-cap: "Spearman Correlation Matrix of Key Numerical Variables"
#| code-fold: true
#| code-summary: "Show Code"
#| fig-width: 10
#| fig-height: 10

cor_vars <- ins_clean %>%
  select(driver_age, vehicle_age, vehicle_value,
         power_to_weight_ratio, seats,
         total_premium, total_incurred, total_claims,
         total_exposure, loss_ratio, claim_frequency) %>%
  drop_na()

cor_matrix <- cor(cor_vars, method = "spearman")

corrplot(
  cor_matrix,
  method      = "color",
  type        = "upper",
  tl.col      = "black",
  tl.srt      = 45,
  tl.cex      = 0.85,
  addCoef.col = "black",
  number.cex  = 0.7,
  col = colorRampPalette(c("#E15759", "white", "#4E79A7"))(200),
  title       = "Spearman Correlation Matrix",
  mar         = c(0, 0, 2, 0)
)
```

### Claim Incidence by Business Type and Payment Frequency

```{r}
#| label: vis18-claim-incidence
#| fig-cap: "Claim Incidence Rate by Business Type and Payment Frequency"
#| code-fold: true
#| code-summary: "Show Code"

claim_rate_data <- ins_clean %>%
  filter(!is.na(business_label), !is.na(payment_frequency)) %>%
  group_by(business_label, payment_frequency) %>%
  summarise(
    n_policies = n(),
    claim_rate = mean(total_claims > 0, na.rm = TRUE),
    .groups    = "drop"
  ) %>%
  mutate(
    se  = sqrt(claim_rate * (1 - claim_rate) / n_policies),
    lower = claim_rate - 1.96 * se,
    upper = claim_rate + 1.96 * se
  )

ggplot(claim_rate_data,
       aes(x = payment_frequency, y = claim_rate,
           fill = business_label)) +
  geom_col(position = "dodge", colour = "white", alpha = 0.85) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    position = position_dodge(0.9), width = 0.2
  ) +
  scale_fill_manual(
    values = c("New Business" = "#4E79A7", "Renewal" = "#59A14F"),
    name   = "Business Type"
  ) +
  scale_y_continuous(labels = percent_format()) +
  labs(title    = "Claim Incidence Rate by Business Type and Payment Frequency",
       subtitle = "Error bars = 95% confidence intervals; non-overlapping bars indicate significant differences",
       x = "Payment Frequency", y = "Proportion of Policies with a Claim") +
  theme_ins
```







