---
title: "1. Visual Analytics for Insurance Portfolio Data Discovery"
author: "Lei Yuhe"
date: "May 17, 2026"
date-modified: "last-modified"
format: html
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

# 1 Overview 

Insurance profitability and volatility are central concerns for non-life insurers. A motor vehicle insurance portfolio accumulates risk across thousands of policyholders, and the underlying drivers of loss, such as vehicle type, driver age, geographic region, claim history are often opaque without analytical investigations. 

As highlighted by [Swiss Re](https://www.swissre.com/dam/jcr:3257ef4b-fe20-45e2-9700-7677702407af/sigma4_2019_en.pdf): *"Insurers often carry underperforming portfolios where the root causes of poor profitability and high volatility are unknown."* Visual analytics offers a powerful lens to investigate these patterns systematically, combining the strengths of statistical methods with human pattern recognition.

This report applies visual analytics techniques to a **real-world Spanish motor vehicle insurance portfolio** dataset from Mendeley Data, to uncover the distribution characteristics of key variables and explore factors influencing portfolio profitability and volatility.

The objective of the analysis is to:

- Understand the overall structure and characteristics of the insurance portfolio.
- Identify profitability and risk patterns across customer segments.
- Detect claim volatility and high-risk profiles.
- Generate actionable business insights for underwriting, pricing, and portfolio management.

# 2 Data Understanding & Preparation

## 2.1 Loading Libraries

::: {.callout-note collapse="true"}
## R Libraries Used

| Library | Purpose |
|---|---|
| `tidyverse` | Data wrangling (`dplyr`, `tidyr`) and base visualisation (`ggplot2`) |
| `ggdist` | Visualising distributions and uncertainty (half-eye plots, dot plots) |
| `ggridges` | Density ridge plots for comparing distributions across groups |
| `ggstatsplot` | Statistical visualisations with built-in hypothesis testing |
| `ggmosaic` | Mosaic plots for categorical variable relationships |
| `patchwork` | Composing multiple ggplot objects into one figure |
| `scales` | Formatting axis labels (percentages, currency) |
| `knitr` | Rendering tables neatly within the report |
| `corrplot` | Correlation matrix visualisation |
| `forcats` | Factor reordering for cleaner categorical plots |

:::

```{r}
#| label: load-libraries
#| code-fold: true
#| code-summary: "Show Code"

pacman::p_load(
  tidyverse,
  ggdist,
  ggridges,
  ggstatsplot,
  ggmosaic,
  patchwork,
  scales,
  knitr,
  corrplot,
  forcats,
  ggrepel,
  ggthemes
)
```

## 2.2 Loading the Data

The dataset is from [Mendeley Data](https://data.mendeley.com/datasets/sw4jmdb2sm/1). It contains policy-level records from a Spanish non-life motor vehicle insurer.

```{r}
#| label: load-data

insurance <- read_delim("data/Dataset of motor insurance portfolio.csv", delim = ";")

```

## 2.3 Data Dictionary

::: {.callout-note collapse="true"}
## Key Variables

| Variable | Type | Description |
|---|---|---|
| `insured_id` | Identifier | Unique policyholder identifier |
| `year` | Numerical | Policy year (2022–2024) |
| `policy_type` | Categorical | Coverage type (COMP_E, COMP_N, CC, TPG, etc.) |
| `policy_status` | Categorical | Policy status (A = Active, C = Cancelled) |
| `business_type` | Categorical | New business (NB) or renewal (P) |
| `payment_frequency` | Categorical | Payment frequency (A = Annual, S = Semi-annual, etc.) |
| `bonus_score` | Categorical | Bonus-malus score reflecting claim history |
| `driver_age` | Numerical | Age of the main driver (years) |
| `vehicle_age` | Numerical | Age of the insured vehicle (years) |
| `age_driving_licence` | Numerical | Years since driving licence obtained |
| `fuel_type` | Categorical | Fuel type (G = Gasoline, D = Diesel, etc.) |
| `vehicle_value` | Numerical | Market value of insured vehicle (EUR) |
| `seats` | Numerical | Number of seats in the vehicle |
| `power_to_weight_ratio` | Numerical | Engine power-to-weight ratio |
| `vehicle_brand` | Categorical | Vehicle manufacturer |
| `municipality_type` | Categorical | Urban (I) or rural (C) classification |
| `circulation_area` | Categorical | Urban (U) or rural (R) circulation area |
| `total_premium` | Numerical | Total gross written premium (EUR) |
| `total_claims` | Numerical | Total number of claims across all coverage types |
| `total_incurred` | Numerical | Total incurred claim cost (EUR) |
| `total_exposure` | Numerical | Policy exposure (fraction of year in force) |

::: 

## 2.4 Deriving Key Variables

Before visualisation, two target variables were derived:

- **Loss Ratio** = `claimed amount / premium`, which measures profitability. A ratio > 1 indicates the insurer is paying out more than it earns.
- **Claim Frequency** = `number of claims / exposure`, which measures volatility by normalising claims to exposure.

```{r}
#| label: feature-engineering
#| code-fold: true
#| code-summary: "Show Code"

insurance <- insurance %>%
  mutate(
    # Core profitability metric
    loss_ratio = total_incurred / total_premium,

    # Core volatility metric
    claim_frequency = total_claims / total_exposure,

    # Whether the policy had at least one claim
    had_claim = if_else(total_claims > 0, "With Claim", "No Claim"),

    # Driver age bands
    driver_age_group = cut(
      driver_age,
      breaks = c(17, 25, 35, 45, 55, 65, Inf),
      labels = c("18-25", "26-35", "36-45", "46-55", "56-65", "65+"),
      right  = TRUE
    ),

    # Vehicle age bands
    vehicle_age_group = cut(
      vehicle_age,
      breaks = c(-1, 2, 5, 10, 15, Inf),
      labels = c("0-2 yrs", "3-5 yrs", "6-10 yrs", "11-15 yrs", "15+ yrs"),
      right  = TRUE
    ),

    # Simplify circulation area labels
    area_label = case_when(
      circulation_area == "U" ~ "Urban",
      circulation_area == "R" ~ "Rural",
      TRUE ~ circulation_area
    ),

    # Simplify fuel type labels
    fuel_label = case_when(
      fuel_type == "G" ~ "Gasoline",
      fuel_type == "D" ~ "Diesel",
      fuel_type == "E" ~ "Electric",
      fuel_type == "H" ~ "Hybrid",
      TRUE ~ fuel_type
    ),

    # Simplify business type labels
    business_label = case_when(
      business_type == "NB" ~ "New Business",
      business_type == "P"  ~ "Renewal",
      TRUE ~ business_type
    )
  )
```

## 2.5 Data Quality Check

```{r}
#| label: data-quality
#| code-fold: true
#| code-summary: "Show Code"

# Missing values
missing_summary <- insurance %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(),
               names_to  = "Variable",
               values_to = "Missing_Count") %>%
  filter(Missing_Count > 0) %>%
  arrange(desc(Missing_Count))

if (nrow(missing_summary) == 0) {
  cat("No missing values detected.\n")
} else {
  kable(missing_summary, caption = "Variables with Missing Values")
}

# Summary of core numerical variables
core_nums <- insurance %>%
  select(driver_age, vehicle_age, vehicle_value, total_premium,
         total_incurred, total_claims, total_exposure,
         loss_ratio, claim_frequency)

kable(summary(core_nums), caption = "Summary Statistics: Core Numerical Variables")
```

## 2.6 Outlier Treatment

Insurance datasets are naturally highly skewed because a small number of policies generate extremely large claims. Without treatment, these extreme cases can dominate the charts and hide meaningful portfolio-level patterns.

Using the 99th percentile filtering improves readability and allows the analysis to focus on the majority of the portfolio while still acknowledging catastrophic claim behaviour.

```{r}
#| label: outlier-handling
#| code-fold: true
#| code-summary: "Show Code"

lr_upper <- quantile(insurance$loss_ratio,      0.99, na.rm = TRUE)
cf_upper <- quantile(insurance$claim_frequency, 0.99, na.rm = TRUE)

cat("Loss Ratio 99th percentile:", round(lr_upper, 4), "\n")
cat("Claim Frequency 99th percentile:", round(cf_upper, 4), "\n")

insurance <- insurance %>%
  mutate(
    is_extreme_lr = loss_ratio      > lr_upper,
    is_extreme_cf = claim_frequency > cf_upper
  )

cat("Policies with extreme LR (> 99th pct):", sum(insurance$is_extreme_lr, na.rm = TRUE), "\n")
cat("Policies with extreme CF (> 99th pct):", sum(insurance$is_extreme_cf, na.rm = TRUE), "\n")

# Clean dataset excludes extremes for primary visualisation
ins_clean <- insurance %>%
  filter(!is_extreme_lr, !is_extreme_cf)
```

---


# 3 Univariate Visual Analytics

## 3.1 Setting a Common Theme

```{r}
#| label: custom-theme
#| code-fold: true
#| code-summary: "Show Code"

theme_ins <- theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, colour = "grey40"),
    axis.title = element_text(size = 10),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

# Colour palettes
policy_colours <- c(
  "COMP_E" = "#4E79A7", "COMP_N" = "#F28E2B",
  "CC" = "#59A14F", "TPG" = "#E15759", "TP" = "#76B7B2"
)

fuel_colours <- c(
  "Gasoline" = "#4E79A7", "Diesel" = "#F28E2B",
  "Electric" = "#59A14F", "Hybrid" = "#E15759"
)
```


## 3.2 Numerical Variables

### 3.2.1 Total Premium Distribution

```{r}
#| label: vis1-premium
#| fig-cap: "Distribution of Total Gross Written Premium"
#| code-fold: true
#| code-summary: "Show Code"

p_prem_hist <- ggplot(ins_clean, aes(x = total_premium)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 60, fill = "#4E79A7",
                 colour = "white", alpha = 0.85) +
  geom_density(colour = "#E15759", linewidth = 1) +
  scale_x_continuous(labels = label_dollar(prefix = "€")) +
  labs(title    = "Distribution of Total Premium",
       subtitle = "Right-skewed; most policies priced under €1,000",
       x = "Total Premium (EUR)", y = "Density") +
  theme_ins

p_prem_box <- ggplot(ins_clean, aes(y = total_premium, x = "")) +
  geom_violin(fill = "#4E79A7", alpha = 0.5) +
  geom_boxplot(width = 0.12, fill = "white", outlier.shape = NA) +
  scale_y_continuous(labels = label_dollar(prefix = "€")) +
  coord_flip() +
  labs(x = NULL, y = "Total Premium (EUR)", title = "Spread") +
  theme_ins

p_prem_hist / p_prem_box + plot_layout(heights = c(3, 1))
```

The premium distribution is heavily right-skewed. The majority of policies carry premiums below €1,000, consistent with standard personal motor policies. The long right tail reflects high-value vehicles and comprehensive coverage tiers. The violin plot confirms the concentration of mass at lower premium bands with a thin upper tail.

### 3.2.2 Total Incurred Claim Cost (Claimants Only)

```{r}
#| label: vis2-incurred
#| fig-cap: "Distribution of Total Incurred Claim Cost"
#| code-fold: true
#| code-summary: "Show Code"

ins_claimants <- ins_clean %>% filter(total_claims > 0)

p_inc_raw <- ggplot(ins_claimants, aes(x = total_incurred)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 60, fill = "#E15759",
                 colour = "white", alpha = 0.85) +
  geom_density(colour = "#4E79A7", linewidth = 1) +
  scale_x_continuous(labels = label_dollar(prefix = "€")) +
  labs(title    = "Raw Total Incurred (Claimants Only)",
       subtitle = "Extreme right skew; most claims are small",
       x = "Total Incurred (EUR)", y = "Density") +
  theme_ins

p_inc_log <- ggplot(ins_claimants, aes(x = log1p(total_incurred))) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 50, fill = "#F28E2B",
                 colour = "white", alpha = 0.85) +
  geom_density(colour = "#4E79A7", linewidth = 1) +
  labs(title    = "Log-Transformed Total Incurred",
       subtitle = "Log scale reveals approximate log-normal structure",
       x = "log(Total Incurred + 1)", y = "Density") +
  theme_ins

p_inc_raw + p_inc_log
```

Claim costs are heavily right-skewed in raw form, with most claims small and a few catastrophic losses dominating the tail. The log-transformed distribution is approximately normal, confirming the log-normal claim severity assumption common in actuarial modelling. This justifies log-scale approaches when modelling severity.


### 3.2.3 Loss Ratio Distribution

```{r}
#| label: vis3-loss-ratio
#| fig-cap: "Policy-Level Loss Ratio Distribution"
#| code-fold: true
#| code-summary: "Show Code"

ggplot(ins_clean, aes(x = loss_ratio)) +
  geom_histogram(
    aes(y = after_stat(density), fill = after_stat(x) > 1),
    bins = 80, colour = "white", alpha = 0.85
  ) +
  scale_fill_manual(
    values = c("FALSE" = "#59A14F", "TRUE" = "#E15759"),
    labels = c("Profitable (LR < 1)", "Unprofitable (LR ≥ 1)"),
    name   = NULL
  ) +
  geom_vline(xintercept = 1, linetype = "dashed",
             colour = "black", linewidth = 0.9) +
  geom_density(colour = "grey30", linewidth = 0.8) +
  annotate("text", x = 1.05, y = Inf, vjust = 2, hjust = 0,
           label = "Break-even (LR = 1)", size = 3.5, colour = "grey30") +
  scale_x_continuous(labels = percent_format()) +
  labs(title    = "Distribution of Loss Ratio Across Policies",
       subtitle = "Policies with LR > 1 represent net losses to the insurer",
       x = "Loss Ratio", y = "Density") +
  theme_ins
```

The loss ratio distribution reveals the characteristic bimodal structure of insurance portfolios: a large spike at zero (policies with no claims) and a long right tail for claiming policies. The vertical break-even line at LR = 1 separates profitable from unprofitable policies. The proportion of policies above LR = 1 represents the portfolio's unprofitable segment, which can be a key metric for policy underwriting.

### 3.2.4 Driver Age and Vehicle Age Distributions

```{r}
#| label: vis4-age
#| fig-cap: "Driver Age and Vehicle Age Distributions"
#| code-fold: true
#| code-summary: "Show Code"

p_driver <- ggplot(ins_clean, aes(x = driver_age)) +
  geom_histogram(bins = 40, fill = "#4E79A7",
                 colour = "white", alpha = 0.85) +
  labs(title    = "Driver Age Distribution",
       subtitle = "Bell-shaped; centred around middle-aged drivers",
       x = "Driver Age (years)", y = "Count") +
  theme_ins

p_vehicle <- ggplot(ins_clean, aes(x = vehicle_age)) +
  geom_histogram(bins = 30, fill = "#F28E2B",
                 colour = "white", alpha = 0.85) +
  labs(title    = "Vehicle Age Distribution",
       subtitle = "Right-skewed; portfolio skews toward newer vehicles",
       x = "Vehicle Age (years)", y = "Count") +
  theme_ins

p_driver + p_vehicle
```

Driver demographics are relatively stable. Driver age follows a roughly bell-shaped distribution centred around middle-aged drivers (35–50), reflecting the active driving population. Vehicle age is right-skewed with a concentration of newer vehicles, likely because newer car owners are more motivated to purchase comprehensive coverage to protect their assets. 

### 3.2.5 Vehicle Value Distribution

```{r}
#| label: vis5-vehicle-value
#| fig-cap: "Distribution of Vehicle Value"
#| code-fold: true
#| code-summary: "Show Code"

vv_upper <- quantile(ins_clean$vehicle_value, 0.99, na.rm = TRUE)

ggplot(ins_clean %>% filter(vehicle_value <= vv_upper),
       aes(x = vehicle_value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = "#76B7B2", colour = "white", alpha = 0.85) +
  geom_density(colour = "#E15759", linewidth = 1) +
  scale_x_continuous(labels = label_dollar(prefix = "€")) +
  labs(title    = "Vehicle Value Distribution (up to 99th percentile)",
       subtitle = "Wide spread; luxury brands (Porsche, Mercedes) create a long tail",
       x = "Vehicle Value (EUR)", y = "Density") +
  theme_ins
```

Vehicle value is highly heterogeneous. The distribution is positively skewed due to high-value luxury vehicles. Vehicle value is a critical risk factor: higher-value vehicles generate larger repair/replacement claims, yet their premiums may not always scale proportionally, creating potential adverse selection in the high-value segment.


## 3.3 Categorical Variables

### 3.3.1 Policy Type Breakdown

::: {.callout-note}
## Policy Type Definitions
- **TP**: Basic third-party liability coverage only
- **TPG**: Third-party liability with glass coverage  
- **CC**: Third-party liability combined with two or more additional coverages (theft, fire, total loss or glass)
- **COMP_E**: Comprehensive with excess
- **COMP_N**: Comprehensive without excess
:::

```{r}
#| label: vis6-policy-type
#| fig-cap: "Policy Count and Premium Contribution by Policy Type"
#| fig-width: 10
#| code-fold: true
#| code-summary: "Show Code"

p_type_count <- ins_clean %>%
  count(policy_type) %>%
  mutate(policy_type = fct_reorder(policy_type, n)) %>%
  ggplot(aes(x = n, y = policy_type, fill = policy_type)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = comma(n)), hjust = -0.1, size = 3.5) +
  scale_fill_manual(values = policy_colours) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Policy Count by Policy Type",
       x = "Number of Policies", y = NULL) +
  theme_ins

p_type_prem <- ins_clean %>%
  group_by(policy_type) %>%
  summarise(total = sum(total_premium)) %>%
  mutate(
    policy_type = fct_reorder(policy_type, total),
    pct = total / sum(total)
  ) %>%
  ggplot(aes(x = total, y = policy_type, fill = policy_type)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = percent(pct, accuracy = 0.1)), hjust = -0.1, size = 3.5) +
  scale_fill_manual(values = policy_colours) +
  scale_x_continuous(labels = label_dollar(prefix = "€", suffix = "M", scale = 1e-6),
  expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Total Premium by Policy Type",
       x = "Total Premium (EUR Millions)", y = NULL) +
  theme_ins

p_type_count + p_type_prem
```

COMP_E (comprehensive with excess) dominates both policy count and premium contribution, reflecting the product mix of this Spanish insurer. The disparity between count share and premium share across policy types reveals pricing tier differences where comprehensive policies command significantly higher premiums per policy than third-party variants (CC, TPG).

### 3.3.2 Business Type and Payment Frequency

```{r}
#| label: vis7-business-payment
#| fig-cap: "New Business vs Renewal and Payment Frequency Breakdown"
#| fig-width: 10
#| code-fold: true
#| code-summary: "Show Code"

p_biz <- ins_clean %>%
  count(business_label) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(x = "", y = pct, fill = business_label)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = paste0(business_label, "\n", percent(pct, accuracy = 0.1))),
            position = position_stack(vjust = 0.5), size = 3.5) +
  scale_fill_manual(values = c("New Business" = "#4E79A7", "Renewal" = "#59A14F")) +
  coord_flip() +
  labs(title = "New Business vs Renewal Split", x = NULL, y = NULL) +
  theme_ins + theme(legend.position = "none",
                    axis.text = element_blank(),
                    axis.ticks = element_blank())

p_pay <- ins_clean %>%
  count(payment_frequency) %>%
  mutate(
    pct       = n / sum(n),
    pay_label = recode(payment_frequency,
                       "A" = "Annually",
                       "S" = "Semi-annually",
                       "Q" = "Quarterly"),
    pay_label = fct_reorder(pay_label, n)
  ) %>%
  ggplot(aes(x = pct, y = pay_label, fill = pay_label)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = percent(pct, accuracy = 0.1)), hjust = -0.1, size = 3.5) +
  scale_x_continuous(labels = percent_format(),
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_manual(values = c(
    "Annually"      = "#2171B5",
    "Semi-annually" = "#6BAED6",
    "Quarterly"   = "#BDD7E7"
  )) +
  labs(title = "Payment Frequency Breakdown",
       x = "Share of Portfolio", y = NULL) +
  theme_ins

p_biz + p_pay
```

The renewal versus new business split indicates portfolio stability, a high renewal rate suggests strong customer retention and lower adverse selection risk (renewals have an established claim history). Annual payment frequency dominates, which is typical for personal motor insurance where policyholders prefer a single annual transaction.


### 3.3.3 Fuel Type and Circulation Area

```{r}
#| label: vis8-fuel-area
#| fig-cap: "Fuel Type and Circulation Area Composition"
#| fig-width: 8
#| code-fold: true
#| code-summary: "Show Code"

p_fuel <- ins_clean %>%
  count(fuel_label) %>%
  mutate(
    fuel_label = fct_reorder(fuel_label, n),
    pct = n / sum(n)
  ) %>%
  ggplot(aes(x = pct, y = fuel_label, fill = fuel_label)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = percent(pct, accuracy = 0.1)), hjust = -0.1, size = 3.5) +
  scale_fill_manual(values = fuel_colours) +
  scale_x_continuous(labels = percent_format(),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Fuel Type Composition",
       x = "Share of Portfolio", y = NULL) +
  theme_ins

p_area <- ins_clean %>%
  count(area_label) %>%
  mutate(
    area_label = fct_reorder(area_label, n),
    pct = n / sum(n)
  ) %>%
  ggplot(aes(x = pct, y = area_label, fill = area_label)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = percent(pct, accuracy = 0.1)), hjust = -0.1, size = 3.5) +
  scale_fill_manual(values = c("Urban" = "#4E79A7", "Rural" = "#59A14F")) +
  scale_x_continuous(labels = percent_format(),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Circulation Area Composition",
       x = "Share of Portfolio", y = NULL) +
  theme_ins

p_fuel + p_area
```

Diesel and gasoline vehicles dominate the portfolio, reflecting Spain's historical fleet composition. Electric and hybrid vehicles are a small but growing segment, likely underrepresented in this dataset's time period. The urban/rural circulation split is important as urban drivers face higher accident frequency due to traffic density, while rural drivers tend to have higher severity claims due to speed.


# 4 Bivariate Visual Analytics

## 4.1 Profitability Analysis (Loss Ratio)

### 4.1.1 Loss Ratio by Policy Type using Raincloud Plot

```{r}
#| label: vis9-lr-policy-type
#| fig-cap: "Loss Ratio Distribution by Policy Type (Raincloud Plot)"
#| code-fold: true
#| code-summary: "Show Code"

ins_clean %>%
  filter(!is.na(policy_type)) %>%
  mutate(policy_type = fct_reorder(policy_type, loss_ratio, .fun = median)) %>%
  ggplot(aes(x = policy_type, y = loss_ratio, fill = policy_type)) +
  stat_halfeye(
    adjust = 0.5, justification = -0.2,
    .width = 0, point_colour = NA, alpha = 0.7
  ) +
  geom_boxplot(width = 0.15, outlier.shape = NA, colour = "grey30") +
  stat_dots(side = "left", justification = 1.1,
            binwidth = 0.008, dotsize = 1.5) +
  geom_hline(yintercept = 1, linetype = "dashed",
             colour = "red", linewidth = 0.7) +
  scale_fill_manual(values = policy_colours) +
  scale_y_continuous(labels = percent_format()) +
  coord_flip() +
  labs(title    = "Loss Ratio Distribution by Policy Type",
       subtitle = "Red dashed line = break-even (LR = 1); raincloud shows full distribution",
       x = NULL, y = "Loss Ratio") +
  theme_ins +
  theme(legend.position = "none")
```

The raincloud plot reveals substantial differences in underwriting performance across policy types. Comprehensive products such as CC and COMP_E exhibit significantly higher variability and more extreme loss ratios compared to TP and TPG products. Although most policies remain profitable with loss ratios below the break-even threshold, a small number of catastrophic claims generate disproportionately large underwriting losses. Thus, insurance profitability is heavily influenced by tail-risk events. 


### 4.1.2 Statistical Test: Loss Ratio by Policy Type

```{r}
#| label: vis10-anova-lr
#| fig-cap: "One-Way ANOVA of Loss Ratio by Policy Type"
#| code-fold: true
#| code-summary: "Show Code"
#| fig-height: 10

ggbetweenstats(
  data  = ins_clean,
  x     = policy_type,
  y     = loss_ratio,
  type  = "nonparametric",
  p.adjust.method = "bonferroni",
  title = "Loss Ratio by Policy Type: Kruskal-Wallis Test",
  xlab  = "Policy Type",
  ylab  = "Loss Ratio",
  messages = FALSE
)
```

The Kruskal-Wallis test (non-parametric, appropriate for the non-normal loss ratio distribution) with Bonferroni-corrected pairwise comparisons statistically validates whether policy type significantly influences loss ratio. The p-value below 0.05 confirms that policy type is a significant driver of profitability differences, justifying its use as a primary rating variable.

### 4.1.3 Loss Ratio by Driver Age Group using Raincloud Plot

```{r}
#| label: vis11-lr-driver-age
#| fig-cap: "Loss Ratio Distribution by Driver Age Group"
#| code-fold: true
#| code-summary: "Show Code"

ins_clean %>%
  filter(!is.na(driver_age_group)) %>%
  ggplot(aes(x = driver_age_group, y = loss_ratio, fill = driver_age_group)) +
  stat_halfeye(adjust = 0.5, justification = -0.2,
               .width = 0, point_colour = NA, alpha = 0.7) +
  geom_boxplot(width = 0.15, outlier.shape = NA, colour = "grey30") +
  geom_hline(yintercept = 1, linetype = "dashed",
             colour = "red", linewidth = 0.7) +
  scale_fill_brewer(palette = "Spectral") +
  scale_y_continuous(labels = percent_format()) +
  coord_flip() +
  labs(title    = "Loss Ratio by Driver Age Group",
       subtitle = "Younger and older drivers typically exhibit higher, more volatile loss ratios",
       x = "Driver Age Group", y = "Loss Ratio") +
  theme_ins +
  theme(legend.position = "none")
```

Young drivers (18–25) and elderly drivers (65+) exhibit elevated loss ratios compared to middle-aged groups. This is consistent with actuarial research: young drivers lack experience while older drivers may have declining reaction times. 


### 4.1.4 Loss Ratio by Circulation Area and Fuel Type using Heatmap

```{r}
#| label: vis12-heatmap-lr
#| fig-cap: "Mean Loss Ratio: Circulation Area vs Fuel Type"
#| code-fold: true
#| code-summary: "Show Code"

heatmap_data <- ins_clean %>%
  group_by(area_label, fuel_label) %>%
  summarise(
    mean_lr = mean(loss_ratio, na.rm = TRUE),
    n       = n(),
    .groups = "drop"
  ) %>%
  filter(!is.na(area_label), !is.na(fuel_label))

ggplot(heatmap_data,
       aes(x = fuel_label, y = area_label, fill = mean_lr)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(
    aes(label = paste0(percent(mean_lr, accuracy = 0.1),
                       "\n(n=", comma(n), ")")),
    size = 3.2, fontface = "bold"
  ) +
  scale_fill_gradient2(
    low = "#59A14F", mid = "#FFFF99", high = "#E15759",
    midpoint = 1,
    labels   = percent_format(),
    name     = "Mean Loss Ratio"
  ) +
  labs(title    = "Mean Loss Ratio: Circulation Area × Fuel Type",
       subtitle = "Green = profitable (< 100%); Red = unprofitable (> 100%)",
       x = "Fuel Type", y = "Circulation Area") +
  theme_ins
```

Urban diesel vehicles exhibit the highest mean loss ratio at 31.7%, while rural gasoline vehicles show the lowest at 24.3%. Across both fuel categories, urban policies consistently generate higher loss ratios than rural policies, suggesting that traffic density and driving environment significantly influence insurance risk. Diesel vehicles also display higher loss ratios than gasoline vehicles regardless of area, indicating that fuel type may be associated with higher vehicle utilisation or claim severity. Despite these differences, all segments remain profitable with mean loss ratios substantially below the 100% break-even threshold. 

### 4.1.5 Loss Ratio by Business Type

```{r}
#| label: vis13-lr-business
#| fig-cap: "Loss Ratio by Business Type (New vs Renewal): Statistical Test"
#| code-fold: true
#| code-summary: "Show Code"


ggbetweenstats(
  data  = ins_clean %>% filter(!is.na(business_label)),
  x     = business_label,
  y     = loss_ratio,
  type  = "nonparametric",
  title = "Loss Ratio: New Business vs Renewal",
  xlab  = "Business Type",
  ylab  = "Loss Ratio",
  messages = FALSE
)
```

The p-value in this case is approximately zero, showing significant difference between distribution of loss ratios of new business and renewals. However, the difference is small in actual practice since the dataset contains huge amount of observations. 

Median loss ratio is zero for both groups, showing that more than half of policies in both groups generated zero or extremely low claims. 

Similarly, both groups display highly skewed distributions with substantial extreme outliers, reflecting the heavy-tailed nature of insurance claims.

## 4.2 Volatility Analysis (Claim Frequency)

### 4.2.1 Claim Frequency by Fuel Type

```{r}
#| label: vis14-cf-fuel
#| fig-cap: "Claim Frequency by Fuel Type: Statistical Test"
#| code-fold: true
#| code-summary: "Show Code"


ggbetweenstats(
  data            = ins_clean %>% filter(!is.na(fuel_label)),
  x               = fuel_label,
  y               = claim_frequency,
  type            = "nonparametric",
  p.adjust.method = "bonferroni",
  pairwise.display = "significant",
  title           = "Claim Frequency by Fuel Type: Kruskal-Wallis Test",
  subtitle        = "Only significant pairwise comparisons shown (Bonferroni-corrected)",
  xlab            = "Fuel Type",
  ylab            = "Claim Frequency (Claims per Exposure Year)",
  messages        = FALSE
)
```

Diesel vehicles exhibit slightly higher claim frequencies than gasoline vehicles, as indicated by the positive rank-biserial correlation coefficient (0.03). However, the effect size remains very small, suggesting that the practical difference between fuel types is limited despite the large sample size.

Overall, fuel type appears to have only modest explanatory power and should be interpreted alongside other underwriting and demographic risk factors.


### 4.2.2 Vehicle Age vs. Loss Ratio: Risk Profile Bubble Chart

```{r}
#| label: vis15-bubble
#| fig-cap: "Risk Profile by Vehicle Age Group"
#| code-fold: true
#| code-summary: "Show Code"

risk_summary <- ins_clean %>%
  filter(!is.na(vehicle_age_group)) %>%
  group_by(vehicle_age_group) %>%
  summarise(
    n_policies      = n(),
    avg_cf          = mean(claim_frequency, na.rm = TRUE),
    avg_lr          = mean(loss_ratio, na.rm = TRUE),
    pct_with_claim  = mean(total_claims > 0, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(risk_summary,
       aes(x = avg_cf, y = avg_lr,
           size = n_policies, colour = vehicle_age_group)) +
  geom_point(alpha = 0.85) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
  geom_text_repel(aes(label = vehicle_age_group), size = 3.5, show.legend = FALSE) +
  scale_size_continuous(name = "Policy Count", range = c(4, 16)) +
  scale_colour_brewer(palette = "Set2", guide = "none") +
  scale_x_continuous(labels = number_format(accuracy = 0.01)) +
  scale_y_continuous(labels = percent_format()) +
  labs(title    = "Risk Profile by Vehicle Age Group",
       subtitle = "Top-right = high frequency & high loss ratio (worst segment); dashed = break-even",
       x = "Average Claim Frequency",
       y = "Average Loss Ratio") +
  theme_ins
```

The bubble chart reveals risk characteristics across vehicle age groups by simultaneously comparing average claim frequency, loss ratio, and portfolio size.

Newer vehicles aged 0–2 years exhibit the highest average claim frequency, suggesting more active claims behaviour among owners of recently purchased vehicles, although the loss ratio remains at low level. 

In contrast, older vehicles aged 15 years and above display the lowest claim frequency and relatively low loss ratios while representing one of the largest portfolio segments.

### 4.2.3 Loss Ratio by Bonus Score: Ridgeline Plot

```{r}
#| label: vis16-ridgeline-bonus
#| fig-cap: "Loss Ratio Distribution by Bonus Score"
#| code-fold: true
#| code-summary: "Show Code"

bonus_median <- ins_clean %>%
  filter(!is.na(bonus_score)) %>%
  mutate(bonus_label = recode(bonus_score,
                               "B" = "Bad",
                               "N" = "Neutral",
                               "G" = "Good")) %>%
  group_by(bonus_label) %>%
  summarise(med_lr = median(loss_ratio, na.rm = TRUE)) %>%
  arrange(desc(med_lr))

ins_clean %>%
  filter(!is.na(bonus_score)) %>%
  mutate(
    bonus_label = recode(bonus_score,
                         "B" = "Bad",
                         "N" = "Neutral",
                         "G" = "Good"),
    bonus_label = factor(bonus_label, levels = bonus_median$bonus_label)
  ) %>%
  ggplot(aes(x = loss_ratio, y = bonus_label,
             fill = stat(quantile))) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE, quantiles = 4, quantile_lines = TRUE
  ) +
  geom_vline(xintercept = 1, linetype = "dashed",
             colour = "red", linewidth = 0.7) +
  scale_fill_manual(
    name   = "Quartile",
    values = c("#d4f1c5", "#c0c0c0", "#c0c0c0", "#f7b6b6")
  ) +
  scale_x_continuous(labels = percent_format(), limits = c(0, 5)) +
  labs(title    = "Loss Ratio by Bonus Score",
       subtitle = "Ordered by median loss ratio, bonus score reflects past claim experience",
       x = "Loss Ratio", y = "Bonus Score") +
  theme_ins
```

There is substantial overlap between the distributions, indicating that bonus score alone is insufficient to fully explain profitability differences.

### 4.2.4 Correlation Matrix of Numerical Variables

```{r}
#| label: vis17-correlation
#| fig-cap: "Spearman Correlation Matrix of Key Numerical Variables"
#| code-fold: true
#| code-summary: "Show Code"
#| fig-width: 10
#| fig-height: 10

cor_vars <- ins_clean %>%
  select(driver_age, vehicle_age, vehicle_value,
         power_to_weight_ratio, seats,
         total_premium, total_incurred, total_claims,
         total_exposure, loss_ratio, claim_frequency) %>%
  drop_na()

cor_matrix <- cor(cor_vars, method = "spearman")

corrplot(
  cor_matrix,
  method      = "color",
  type        = "upper",
  tl.col      = "black",
  tl.srt      = 45,
  tl.cex      = 0.85,
  addCoef.col = "black",
  number.cex  = 0.7,
  col = colorRampPalette(c("#E15759", "white", "#4E79A7"))(200),
  title       = "Spearman Correlation Matrix",
  mar         = c(0, 0, 2, 0)
)
```

The correlation heatmap reveals that most numerical variables exhibit relatively weak linear relationships. 

Strong positive correlations are observed between driver age and vehicle age (0.86), indicating that older drivers tend to own older vehicles. Loss ratio demonstrates extremely strong relationships with total incurred claims and claim frequency, which shows the mathematical dependence between these variables.

### 4.2.5 Claim Incidence by Business Type and Payment Frequency  

```{r}
#| label: vis18-claim-incidence
#| fig-cap: "Claim Incidence Rate by Business Type and Payment Frequency"
#| code-fold: true
#| code-summary: "Show Code"

claim_rate_data <- ins_clean %>%
  filter(!is.na(business_label), !is.na(payment_frequency)) %>%
  group_by(business_label, payment_frequency) %>%
  summarise(
    n_policies = n(),
    claim_rate = mean(total_claims > 0, na.rm = TRUE),
    .groups    = "drop"
  ) %>%
  mutate(
    se  = sqrt(claim_rate * (1 - claim_rate) / n_policies),
    lower = claim_rate - 1.96 * se,
    upper = claim_rate + 1.96 * se
  )

ggplot(claim_rate_data,
       aes(x = payment_frequency, y = claim_rate,
           fill = business_label)) +
  geom_col(position = "dodge", colour = "white", alpha = 0.85) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    position = position_dodge(0.9), width = 0.2
  ) +
  scale_fill_manual(
    values = c("New Business" = "#4E79A7", "Renewal" = "#59A14F"),
    name   = "Business Type"
  ) +
  scale_y_continuous(labels = percent_format()) +
  labs(title    = "Claim Incidence Rate by Business Type and Payment Frequency",
       subtitle = "Error bars = 95% confidence intervals, non-overlapping bars indicate significant differences",
       x = "Payment Frequency", y = "Proportion of Policies with a Claim") +
  theme_ins
```

Renewal policies consistently exhibit higher proportions of policies with claims compared to new business policies across all payment frequencies.

Quarterly payment customers demonstrate the highest claim incidence rates, while annual payment customers display the lowest.

Although renewal customers exhibit higher claim incidence, earlier analyses in the risk profile bubble chart showed relatively stable profitability levels, implying that higher claim frequency does not necessarily correspond to higher claim severity.





